---
title: "Managing projects with R"
author: "Instructors: Daniel Falster & Will Cornwell"
date: "17/12/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Welcome

Logistics

- nametags: sticker and desk
- post it  notes, 

Getting to know you:

- who has used R
- best & worst experience, worries, why here

Aims of this course:

- Introduction to R & Rstudio
- Getting data in and out of R 
- R objects and classes
- Packages
- Data manipulation
- Data visualisation

Brief recap from yesterday

- R & Rstudio
- packages
- loading data
- assigning variables
- object types 
- Rmd files

What is the tidyverse

* The [Tidyverse](http://tidyverse.org) is the name given to suite of R packages designed for seamless data analysis
* Designed to help you fall into a **"Pit of Success"**
* Tools designed for: 1) Turning data into tidy data, and 2) Plotting & analysing Tidy Data
* Dataframes (tibbles) are the universal "tidy" input and output
* A tidy dataframe has: 1 observation per row, 1 variable per column
* Not one but a collection packages

Can load individually or all together

```{r, eval = FALSE}
library(dplyr)
library(readr)
library(tidyr)
# ...
```
or
```{r, results='hide',message=FALSE}
library(tidyverse)
```

# Project management

(Dan, 9:30-10:30)

## Motivation

Live is complicated, and I used to be in pain, but now I live a blissful peaceful happy existence with no worries aside from feeding unicorns.

**Exercises:**

Challenge - path names:

1. Use the `dir` function to list contents of a folder

Challenge - projects:

1. Create an organized project directory for this stats course
2. Include an RStudio Project in the folder
3. Download the course data (from Moodle or github)
4. Create a file something like 'R_day1.R' 
5. Write one line of code to make a variable, and then save the object as a .rds file
6. When you finish put the green post-it on your laptop screen
7. Put up your green sticky when you're done
  - don't forget to use your directories (/data, /R, /output)
  
Optional:

Organize one of your own projects
  
# Reproducible research with Rmarkdown

(Dan, 10:30-11:00)


Links: 

- blogdown: 
- Pimp my Rmd: https://holtzy.github.io/Pimp-my-rmd/

# Morning tea

(11:00-11:15)

