---
title: "Water quality at Sydney City Beaches"
author: "Instructors: Daniel Falster & Will Cornwell"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(scales)
library(ggbeeswarm)
library(lubridate)
```

Data from OEH Beach water quality data, available [here](https://www.environment.nsw.gov.au/topics/water/beaches/beach-water-quality-data). Data for all of NSW can be downloaded at [the download tab](https://www.environment.nsw.gov.au/beachapp/report_enterococci.aspx). Thanks for Jenny Richmond from [RLadies Sydney](https://rladiessydney.org/) for introducing this great resource (see Jenny's write-up [here](https://rladiessydney.org/post/2018/11/05/basicbasics-3/))

We have extracted a dataset for Sydney city, stored in `data/sydneybeaches.csv`.

<!-- 
x <- read_csv( "EnterococciDataDownload-20181218-090215.csv") %>% 
filter(x, Region ==  "Sydney City Ocean Beaches", !is.na(Enterococci_cfu_per_100ml)) %>% write_csv("data/sydneybeaches.csv")
-->

Load data, extract year using the package lubridate

```{r,warning=FALSE, results='hide', message=FALSE}
data <- read_csv("data/sydneybeaches.csv") %>% 
  mutate(Date = Date %>% lubridate::dmy(),
         year = Date %>% lubridate::year()
         ) %>% 
  filter(!is.na(Enterococci_cfu_per_100ml)) 
```


```{r}
sites <- data %>% 
  select(Region, Site, Latitude, Longitude) %>% 
  distinct()
```

```{r}
library(leaflet)

leaflet() %>%
  addTiles() %>%
  addMarkers(lng = sites$Longitude, lat = sites$Latitude,
             label = paste(sites$Site, "(", sites$Region, ")"))	%>%
  addScaleBar() %>%
  addMiniMap(zoomLevelFixed=2)

```

Question: How does the distribution of values change across years and beaches?

Plots using these package ggplot and ggbeeswarm:

```{r, warning=FALSE}
data %>% 
  ggplot(aes(y = Enterococci_cfu_per_100ml, x = year, colour = log(Enterococci_cfu_per_100ml))) +
    ggbeeswarm::geom_quasirandom() +
    scale_y_log10( name="Enterococci (cfu/100ml)") +
    theme(legend.position = "none")+ 
    facet_wrap(vars(Site)) 
```

Check link to rainfall (in Randwick where we have data):

```{r}
rain <- read_csv("data/randwickrain.csv") %>% 
  mutate(
    Date = date %>% lubridate::ymd(),
    rain = 0.01 + rain_mm
    )

data_randwick <- data %>% 
  filter(year > 2012, Council == "Randwick Council") %>% 
  left_join(rain, by="Date")
```

Check against rainfall on that day:

```{r}
data_randwick %>% 
  ggplot(aes( x = rain, y = Enterococci_cfu_per_100ml, colour = log(Enterococci_cfu_per_100ml ))) +
    geom_point() +
    scale_y_log10( name="Enterococci (cfu/100ml)") + 
    scale_x_log10( name="Rainfall (mm)") +
    theme(legend.position = "none") + 
    facet_wrap(vars(Site)) 
```

Split by beach and year:

```{r}
data_randwick %>% 
  filter(Site %in% c( "Maroubra Beach", "Coogee Beach", "Malabar Beach")) %>% 
  ggplot(aes( x = rain, y = Enterococci_cfu_per_100ml, col = Site)) +
    geom_point() +
    scale_y_log10( name="Enterococci (cfu/100ml)")+ 
    scale_x_log10( name="Rainfall (mm)") +
    facet_wrap(vars(year)) 

```


