---
title: "Data Manipulation and Visualization"
author: "Instructors: Daniel Falster & Will Cornwell"
date: "17/12/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Brief recap 

- load data
- dataframes (tibbles)

## Load individually or all together

```{r, eval = FALSE}
library(dplyr)
library(readr)
library(tidyr)
# ...
```
or
```{r, results='hide',message=FALSE}
library(tidyverse)
```

#Dplyr intro (Will 11:15 - 12:15)

- Data is never organized in the way you want it 

- High % of project is data wrangling

- Many many many modern jobs are data wrangling

# data is a noun

# actions on data are verbs  

# learn a few key verbs and what they do

read_csv
filter
select
mutate
arrange

group_by
summarise
count 
n

rename

join family

pipes

Exercise


Lunch 1230-1330

Dpylr excersise 1330-1400

## ggplot philosophy (1400-1500)

- fonti intro talk 

Extra 

- ggplot vs base R
- grammar of graphics
- types of plots (inlc extra packages, ggbeeswarm)
- themes
- special axes (log)
- facets
- colours, symbols
- ggsave
- ?patchwork (load via source install)
- stats: gmsmooth, lm


**Exercises**: 

1.

# Data visualisation

## Generate hypotheses

- Using dataset from 1st half day, discuss in groups a plot you'd like to make
- Report idea to group

Sydney beaches data



## Exercises


(revise) Challenge - basic ggplot:

1. Load a new dataset
2. Choose a plot type, and plot something
3. Set a colour/fill/shape aesthetic
4. Choose a theme
5. Choose a facet if applicable, or scale one of your aesthetics
6. Assign the plot to an object
7. Save the plot as a couple different file formats



