---
title: "Data Manipulation and Visualization"
author: "Instructors: Daniel Falster, Will Cornwell, Dony Indiarto"
date: "Feb 2020"
output: html_document
---

# Introduction to using Rmarkdown (Aniko)

- File extension `*.Rmd`
- Can be notebook or markdown
- Includes text and code -> make it good for note taking and reports 

Code chunks:

- run standard R code
- are embedded in a wrapper like this

```{r}

```

Text is written in [markdown](https://en.wikipedia.org/wiki/Markdown) --  a lightweight markup language with plain-text-formatting syntax.

- Headers: `#`
- Bold: `**bold**`
- links: `[text](link)`

The Rstudio interface makes it easy to work with both Rmarkdown (`*.Rmd`) and markdown files (`*.md`):

- easily run code chunks
- easily navigate
- use the document outline in the top corner to jump between headers

**Exercise**:  Create a new R markdown document (File -> New -> R Markdown, choose HTML). Add some text, then some code. Run the code chunks. Compile into an html file.   

# Introduction to movies dataset (Fonti)

For today's exercises we're using a data on movie production, from the ["Tidy Tuesday" collection](https://github.com/rfordatascience/tidytuesday). 

```{r}
library(tidyverse)
movies <- read_csv("data/movie_profit/movie_profit.csv")
```

What variables?
```{r}
names(movies)
```

An overview of the data

```{r}
#View(movies)
```

or summarise using the `skimr` package:

```{r}
skimr::skim(movies)
```

**Exercise**: With your partner, review the `README` for the data at `data/movie_profit/readme.md` to see the variables included. 


# What is the tidyverse? (Will)

* The [Tidyverse](http://tidyverse.org) is the name given to suite of R packages designed for seamless data analysis
* Designed to help you fall into a **"Pit of Success"**
* Tools are designed to work seamlessly together, for: 1) Turning data into tidy data, and 2) Plotting & analysing Tidy Data
* Not one but a collection packages
* Dataframes (tibbles) are the universal "tidy" input and output


Load (and install) individually or all together

```{r, eval = FALSE}
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
```
or
```{r, eval = FALSE}
library(tidyverse)
```

**Exercise**: Make sure you can load the `tidyverse`.


## What is a `tibble`? 

Basically, it's a better data frame.

Key features of a dataframe:

- table of data
- each column may have a different data type: character, double, integer, bool, string.

Tibbles have extra functionality:

* Prints a short summary at console
* Generates suitable warnings 
* Won't convert strings to factors
* No partial column-name matching
* Behaves consistently with `[` and `[[`
* For more, see [tibble.tidyverse.org](tibble.tidyverse.org)

```{r}
movies
```

# Reading data (Aniko)

Where possible, we recommend using 

- `csv` files (common separated files) to store data
- the `readr` package to read and write data

Why `readr`?

* Fast
* Won't convert strings to factors
* Can control how columns are parsed

```{r}
movies <- read_csv("data/movie_profit/movie_profit.csv")
```

Has a bunch of extra arguments: guess_max, trim_ws, skip, na

**Exercises**: 

1. Review the help for `read_csv` and figure out what one of the above arguments does.
2. Use `read_csv` to load data file `data/plant_height/plantHeightSingleSpp.csv`
3. Use `read_csv` to load data file `data/sydney_beaches/temperature.csv`

When reading data, `read_csv` tries to guess column types. We can override this by specifying column types exactly.

Read everything as characters

```{r}
read_csv("data/movie_profit/movie_profit.csv", 
         col_types = cols(.default = col_character())
)
```

Specify by variable-by-variable:

```{r}
read_csv("data/movie_profit/movie_profit.csv", 
         col_types = cols(
           X1 = col_integer(),
           release_date = col_character(),
           movie = col_character(),
           production_budget = col_double(),
           domestic_gross = col_double(),
           worldwide_gross = col_double(),
           distributor = col_character(),
           mpaa_rating = col_character(),
           genre = col_character()
         )
)
```

Also other variants

- `read_tsv`: load tab-separated data
- `read_delim`: load file by specifying the delimiter

```{r}
read_delim("data/movie_profit/movie_profit.csv", ",")
```

To load data from excel spreadsheets, try the `readxl` package.

# Data manipulation with `dplyr` (Dan)

Motivation:

- Data is never organized in the way you want it 
- High % of project is data wrangling
- Many many many modern jobs are data wrangling

**Exercise**: Together with your partner, come up with 3 types of change you may need to make on a dataset before it is ready for analysis.




`dplyr` used verbs to describe the actions we want to take on the data

- `select`   -> subset columns 
- `filter`   -> subset rows 
- `arrange`  –> order rows
- `rename`   –> rename variables
- `mutate`   –> make new variables
- `summarise`–> summarise data
- `distinct` -> filter to each unique row

Examples:

To select certain variables:

```{r}
#select(movies, ???)
```

To filter to particular rows:

```{r}
filter(movies, distributor == "Universal")
```

To sort by certain variables:

```{r}
arrange(movies, distributor, production_budget)
```

## Pipes

The pipe is a bit of magic. It's written by `%>%` (Shift-Command-M on Mac or Shift-Control-M on PC ). We can use "the pipe" [%>%](http://magrittr.tidyverse.org/reference/pipe.html) to connect expressions

* `%>%` is an **infix operator** -> expects commands on left & right
* Comes from the [magrittr](http://magrittr.tidyverse.org/reference/pipe.html) package
* `%>%` "pipes" the **output** of the last expression as the **first input** of the next expression
* If you use RStudio, you can type the pipe with Ctrl + Shift + M if you have a PC or Cmd + Shift + M if you have a Mac.

Examples:

```{r}
movies$distributor %>% unique()
```

```{r}
movies$distributor %>% unique() %>% length()
movies$distributor %>% n_distinct()
```

But you can control the input position of the next function with `.`:
```{r}
20 %>% seq(1, 4, length.out = .)
```


Tidyverse functions are written to work with pipes, i.e. most take the data as the first argument.

```{r}
filter(movies, distributor == "Universal")
```

is the same as

```{r}
movies %>% filter(distributor == "Universal")
```

This means we can use pipes to join data verbs to make a data sentence. 

```{r}
movies %>%
  filter(distributor == "Universal") %>%
  select(movie, mpaa_rating)
```

**Exercises:** Apply the `dplyr` package and your new data wrangling skills to the movies dataset to 


1. Create a subset of the dataset consisting of only movies distributed by `Walt Disney`
2. As above and only including variables `movie`, `worldwide_gross`, and `production_budget`
3. As above but with data sorted alphabetically by `movie`
4.  As above but with an additional column return given by `worldwide_gross/production_budget`


```{r}

```

# Imagine your plot (Will)

https://www.data-to-viz.com


# Intro to data visualisation with `ggplot2` (Fonti)

See slides in "slides/datavis_with_ggplot2.html"

**Examples from the slides:**

Read and inspect the data

```{r}
library(tidyverse)
movies <- read_csv("data/movie_profit/movie_profit.csv")
glimpse(movies) # see every column in the data frame
```

Q1 What are the most common genres? 

```{r}
ggplot(movies,         # the data
       aes(x= genre))+ # what to plot
  geom_bar()           # type of plot

```


Q1 And put some colour on...

```{r }
ggplot(movies,                        # the data
       aes(x = genre, fill = genre))+ # what to plot
  geom_bar()                          # type of plot

```

Q2 What genres are more expensive to produce?

```{r}
ggplot(movies, aes(x = genre, y = production_budget))+
  geom_boxplot()
```


Q3 What genres make the highest profit?

```{r fig.height=3, message=FALSE, warning=FALSE, dev='svg', highlight.output=c(1)}
ggplot(movies, aes(x = genre, y = worldwide_gross))+
   geom_boxplot()
```

Q3 re-scale so we can see more variation...

```{r fig.height=3, message=FALSE, warning=FALSE, dev='svg', highlight.output=c(3)}
ggplot(movies,aes(x = genre, y = worldwide_gross))+
  geom_boxplot() +
  scale_y_log10()
```


**Exercises**: Try to answer these questions below by making some relevant figures

1. Are there more cheap films?

 - Hint: create a histogram-plot with production_budget at the x-axis

```{r}

```
 
   
2. What genres make the highest return on investment?

 - Hints
   - calculate the ratio of worldwide gross profit and production budget
   - create a box-plot with x-axis representing the genre and y-axis representing the ratio
   - Rescale y-axis to log scale

```{r}

```
   
3. What genres make the highest return on investment?

 - Hints
   - create a scatter plot with with the x-axis representing the production_budget and the y-axis representing the worldwide_gross profit
   - Rescale the x- and y-axis to log scale
   - add regression line with `stat_smooth()`

```{r}

```

## What to do when your big pipe thingo doesn't work?
