---
title: "Data Manipulation and Visualization"
author: "Instructors: Daniel Falster, Will Cornwell, Fonti Kar, Anikó Tóth, Dony Indiarto"
date: "Feb 2021"
output: html_document
---

# Introduction to using Rmarkdown (Aniko)

- File extension `*.Rmd`
- Can be notebook or markdown
- Includes text and code -> make it good for note taking and reports 

Code chunks:

- run standard R code
- are embedded in a wrapper like this

```{r}

```

Text is written in [markdown](https://en.wikipedia.org/wiki/Markdown) --  a lightweight markup language with plain-text-formatting syntax.

- Headers: `#`
- Bold: `**bold**`
- links: `[text](link)`

The Rstudio interface makes it easy to work with both Rmarkdown (`*.Rmd`) and markdown files (`*.md`):

- easily run code chunks
- easily navigate
- use the document outline in the top corner to jump between headers

**Exercise**:  Create a new R markdown document (File -> New -> R Markdown, choose HTML). Add some text, then some code. Run the code chunks. Compile into an html file.   

# Introduction to movies dataset (Fonti)

For today's exercises we're using a data on movie production, from the ["Tidy Tuesday" collection](https://github.com/rfordatascience/tidytuesday). 

```{r}
library(tidyverse)
movies <- read_csv("data/movie_profit/movie_profit.csv")
```

What variables?
```{r}
names(movies)
```

An overview of the data

```{r}
#View(movies)
```

or summarise using the `skimr` package:

```{r}
skimr::skim(movies)
str(movies)
```

**Exercise**: With your partner, review the `README` for the data at `data/movie_profit/readme.md` to see the variables included. 


# What is the tidyverse? (Will)

* The [Tidyverse](http://tidyverse.org) is the name given to suite of R packages designed for seamless data analysis
* Designed to help you fall into a **"Pit of Success"**
* Tools are designed to work seamlessly together, for: 1) Turning data into tidy data, and 2) Plotting & analysing Tidy Data
* Not one but a collection packages
* Dataframes (tibbles) are the universal "tidy" input and output


Load (and install) individually or all together

```{r, eval = FALSE}
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
```
or
```{r, eval = FALSE}
library(tidyverse)
```

**Exercise**: Make sure you can load the `tidyverse`.

```{r}
library(tidyverse)
```



## What is a `tibble`? 

Basically, it's a better data frame.

Key features of a dataframe:

- table of data
- each column may have a different data type: character, double, integer, bool, string.

Tibbles have extra functionality:

* Prints a short summary at console
* Generates suitable warnings 
* Won't convert strings to factors
* No partial column-name matching
* Behaves consistently with `[` and `[[`
* For more, see [tibble.tidyverse.org](tibble.tidyverse.org)

```{r}
movies
```

# Reading data (Aniko)

Where possible, we recommend using 

- `csv` files (common separated files) to store data

```{r}
movies <- read.csv("data/movie_profit/movie_profit.csv")
```

Has a bunch of extra arguments: skip, check.names

**Exercises**: 

1. Review the help for `read.csv` and figure out what one of the above arguments does.
2. Use `read.csv` to load data file `data/plant_height/plantHeightSingleSpp.csv`
3. Use `read.csv` to load data file `data/sydney_beaches/temperature.csv`

```{r}
#2. Use `read.csv` to load data file `data/plant_height/plantHeightSingleSpp.csv`
read.csv("data/plant_height/plantHeightSingleSpp.csv")
```

```{r}
#3. Use `read.csv` to load data file `data/sydney_beaches/temperature.csv`
read.csv("data/sydney_beaches/temperature.csv")
```


When reading data, `read.csv` tries to guess column types. We can override this by specifying column types exactly.

Read everything as characters

```{r}
read.csv("data/movie_profit/movie_profit.csv", colClasses = "character")
```

Also other variants

- `read.delim`: load file by specifying the delimiter

```{r}
read.delim("data/movie_profit/movie_profit.csv", sep=",")
```

To load data from excel spreadsheets, try the `readxl` package.

```{r}
Example here 
```


# Data manipulation with `dplyr` (Dan)

Motivation:

- Data is never organized in the way you want it 
- High % of project is data wrangling
- Many many many modern jobs are data wrangling

**Exercise**: Together with your partner, come up with 3 types of change you may need to make on a dataset before it is ready for analysis.




`dplyr` used verbs to describe the actions we want to take on the data

- `select`   -> subset columns 
- `filter`   -> subset rows 
- `arrange`  –> order rows
- `rename`   –> rename variables
- `mutate`   –> make new variables
- `summarise`–> summarise data
- `distinct` -> filter to each unique row

Examples:

To select certain variables:

```{r}
select(movies, genre)
```

To filter to particular rows:

```{r}
filter(movies, distributor == "Universal")
```

To sort by certain variables:

```{r}
arrange(movies, distributor, production_budget)
```

## Pipes

The pipe is a bit of magic. It's written by `%>%` (Shift-Command-M on Mac or Shift-Control-M on PC ). We can use "the pipe" [%>%](http://magrittr.tidyverse.org/reference/pipe.html) to connect expressions

* `%>%` is an **infix operator** -> expects commands on left & right
* Comes from the [magrittr](http://magrittr.tidyverse.org/reference/pipe.html) package
* `%>%` "pipes" the **output** of the last expression as the **first input** of the next expression
* If you use RStudio, you can type the pipe with Ctrl + Shift + M if you have a PC or Cmd + Shift + M if you have a Mac.

Examples:

```{r}
movies$distributor %>% unique()
```

```{r}
movies$distributor %>% unique() %>% length()
movies$distributor %>% n_distinct()
```

But you can control the input position of the next function with `.`:
```{r}
20 %>% seq(1, 4, length.out = .)
```


Tidyverse functions are written to work with pipes, i.e. most take the data as the first argument.

```{r}
filter(movies, distributor == "Universal")
```

is the same as

```{r}
movies %>% filter(distributor == "Universal")
```

This means we can use pipes to join data verbs to make a data sentence. 

```{r}
movies %>%
  filter(distributor == "Universal") %>%
  select(movie, mpaa_rating)
```

**Exercises:** Apply the `dplyr` package and your new data wrangling skills to the movies dataset to 


1. Create a subset of the dataset consisting of only movies distributed by `Walt Disney`
2. As above and only including variables `movie`, `worldwide_gross`, and `production_budget`
3. As above but with data sorted alphabetically by `movie`
4. As above but with an additional column return given by `worldwide_gross/production_budget`


```{r}
# . Create a subset of the dataset consisting of only movies distributed by `Atlantic`
m1 <- movies %>% 
   filter(distributor=="Lionsgate")
   
# 2. As above and only including variables `movie`, `worldwide_gross`, and `production_budget`
m2 <- m1 %>% 
   select('movie', 'worldwide_gross', 'production_budget')

# 3. As above but with data sorted alphabetically by `movie`
m3 <- m2 %>% 
   arrange(movie)

# 4.  As above but with an additional column return given by `worldwide_gross/production_budget`
m3 %>% 
   mutate(ratio = worldwide_gross/production_budget)

```

# Imagine your plot (Will)

https://www.data-to-viz.com

# Intro to data visualisation with `ggplot2` (Fonti)

See slides in "slides/datavis_with_ggplot2.html"

**Examples from the slides:**

## An example

```{r}
ggplot(data = movies, aes(x = genre, y = production_budget)) + 
  geom_boxplot() 
```

## More data exploration plots
```{r}
#Density/Histograms
ggplot(data = movies, aes(x = log(domestic_gross))) + 
  geom_histogram(aes(y = ..density..)) +
  geom_density() 

#Scatter plots for exploring relationships
ggplot(data = movies, aes(x = production_budget, y = domestic_gross)) +   geom_point(alpha = 0.5)

data_summary <- function(x) {
   m <- mean(x)
   ymin <- m-sd(x)
   ymax <- m+sd(x)
   return(c(y=m,ymin=ymin,ymax=ymax))
}

#Violin plots with customised function
ggplot(data = movies, aes(x = genre, y = production_budget)) + 
  geom_violin(trim = F) + 
  geom_point(colour = "navy", size = 1, alpha = 0.1) + 
  stat_summary(fun.data="data_summary", col = "black") 
```

## Why plot: Communicate science, results
```{r}
#Barplot production budget by Top 6 most common distributors
top_6 <- movies %>% tabyl(distributor) %>% arrange(-n) %>% 
  filter(n >= 240) %>% pull(distributor)

A <- movies %>% filter(distributor %in% top_6) %>% 
ggplot(aes(x = reorder(distributor, -production_budget, sum), y = production_budget, fill = distributor)) + 
  geom_col() + 
  ylab("Production budget (USD)") + 
  xlab("Distribution companies") + 
  scale_fill_manual(values = c("#A42820", "#5F5647", "#3F5151", "#4E2A1E", "#550307", "#0C1707")) + 
  coord_flip() + 
  theme_bw() + 
  theme(legend.position = "none")
A

#Side by side plot of domestic gross and worldwide gross of Top 6 most common distributors
distributor_long <- movies %>%
  filter(distributor %in% top_6) %>% 
  select(distributor, domestic_gross, worldwide_gross) %>% 
  pivot_longer(cols = 2:3, names_to = "type")

B <- ggplot(distributor_long, aes(x = reorder(distributor, -value, sum), y = value, fill = type)) + 
  geom_col(position = "dodge") +
  ylab("Gross income (USD)") + 
  xlab("Distribution companies") + 
  scale_fill_manual(values = c("#550307", "#0C1707"), name = "Type of income", labels = c("Domestic", "Worldwide")) + 
  coord_flip() + 
  theme_bw() + 
  theme(legend.position = "bottom",
        legend.title = element_text(size = 10),
        axis.title.y = element_blank(),
        axis.text.y = element_blank())
B

#Plot release date by production budget
#Set release date as a date
movies <- movies %>% 
  mutate(release_date = mdy(release_date))

C <- ggplot(data = movies, aes(x = release_date, y = production_budget)) + 
  geom_point(alpha = 0.4, colour = "#0C1707") + 
  xlab("Release Date (Year)") + 
  ylab("Production budget (USD)") + 
  theme_bw() 
C

#Production cost and worldwide gross
D <- ggplot(data = movies, aes(x = production_budget, y = worldwide_gross)) + 
  geom_point(alpha = 0.4, colour = "#0C1707") + 
  geom_point(data = movies %>% filter(production_budget > 150000000 & worldwide_gross > 1000000000), colour = "#A42820") +
  geom_text(data = movies %>% filter(production_budget > 150000000 & worldwide_gross > 1000000000), aes(label = movie), size = 3, hjust = 1.1) + 
  geom_point(data = movies %>% filter(production_budget < 50000000 & worldwide_gross > 787000000), colour = "#5F5647") + 
  geom_text(data = movies %>% filter(production_budget  < 50000000 & worldwide_gross > 787000000), aes(label = movie), size = 3, hjust = -0.1) + 
  geom_point(data = movies %>% filter(production_budget < 100000000 & worldwide_gross > 1150000000), colour = "gold") + 
  geom_text(data = movies %>% filter(production_budget < 100000000 & worldwide_gross > 1150000000), aes(label = movie), size = 3, hjust = -0.2) + 
  ylab("Worldwide gross (USD)") + 
  xlab("Production budget (USD)") + 
  theme_bw() 
D

(A + B + C) / D
```


## Summarise and plot

```{r}
mpaa_rating_means <- movies %>% 
  filter(! is.na(mpaa_rating)) %>% 
  group_by(mpaa_rating) %>% 
  summarise(mean_worldwide_gross = mean(worldwide_gross),
            se_worldwide_gross = sd(worldwide_gross)/sqrt(length(worldwide_gross)),
            lower_worldwide_gross = mean_worldwide_gross - se_worldwide_gross,
            upper_worldwide_gross = mean_worldwide_gross + se_worldwide_gross) 

head(mpaa_rating_means)

ggplot(data = mpaa_rating_means, aes(x = mpaa_rating, y = mean_worldwide_gross)) + 
  geom_col() + 
  geom_errorbar(aes(ymin = lower_worldwide_gross, 
                    ymax = upper_worldwide_gross),
                width = 0.2)
  
```

## Customise "Accessorising"

```{r}
#Colour themes
ggplot(data = movies, aes(x = genre, y = production_budget)) + 
  geom_boxplot() + 
  theme_classic() 

ggplot(data = movies, aes(x = genre, y = production_budget)) + 
  geom_boxplot() + 
  theme_bw() 

ggplot(data = movies, aes(x = genre, y = production_budget)) + 
  geom_boxplot() + 
  theme_dark() 

ggplot(data = movies, aes(x = genre, y = production_budget)) + 
  geom_boxplot() + 
  theme_void() 
```

```{r}
#Axis labels
ggplot(data = movies, aes(x = genre, y = production_budget)) + 
  geom_boxplot() +
  ylab("Production budget (USD)") + 
  xlab("Genre") + 
  theme_bw() 

#Axis labels, colour, points and size/type face/font
ggplot(data = movies, aes(x = genre, y = production_budget, fill = genre)) + 
  geom_boxplot(outlier.colour = "red", size = 1.2) +
  ylab("Production budget (USD)") + 
  xlab("Genre") + 
  theme_bw() + 
  theme(axis.title.x = element_text(family = "Times", face = "bold.italic", size = 12),
        axis.title.y = element_text(family = "Times", face = "bold.italic", size = 12))
```

## Create your own theme - do it once and do it right

```{r}
thesis_theme <- theme_bw() + 
  theme(axis.title.x = element_text(family = "Calibri", face = "bold", size = 16),
        axis.title.y = element_text(family = "Calibri", face = "bold", size = 16),
        axis.text.x = element_text(family = "Calibri" , size = 12),
        axis.text.y = element_text(family = "Calibri", size = 12),
        strip.text = element_text(family = "Calibri", size = 12))

ggplot(data = movies, aes(x = genre, y = production_budget)) + 
  geom_boxplot() +
  ylab("Production budget (USD)") + 
  xlab("Genre") + 
  thesis_theme
```

## A ecostatistician's tool box: 
### Facet wraps
```{r}
top6_plot <- movies %>% filter(distributor %in% top_6 & !is.na(mpaa_rating)) %>% 
  ggplot(data = ., aes(x = production_budget, y = domestic_gross)) +  
  geom_point() +
  theme_bw()

top6_plot + 
  facet_wrap(~distributor) 

top6_plot + 
  facet_grid(distributor ~ genre) 
```

### Scientific labels
https://robjhyndman.com/hyndsight/latex2exp/

```{r}
#Make some fake data
n <- 25
# the paramters
beta_0 <- 0.0005 # the true intercept
beta_1 <- 0.02 # the true slope
sigma <- 0.03 # the true standard deviation

temperature <- seq(20,60, by = 4)
metabolic_rate <- beta_0 + beta_1*temperature + rnorm(temperature, sd=sigma)
fake_data <- data_frame(temperature, metabolic_rate)

ggplot(fake_data, aes(x = temperature, y = metabolic_rate)) + 
  geom_point(size = 3) + 
  scale_y_log10() +
  ylab(TeX("log VCO_2 mL min^{-1}")) + 
  xlab(TeX('Temperature ^{º}C')) + 
  theme(axis.title.y = element_text(size = 30),
        axis.title.x = element_text(size = 30))
```

### Add a bit of flair: colour schemes
-Viridis, great for continuous data
-Rcolourbrewer Colour blind friendly palettes
-Wes Anderson palettes

```{r}
#Colour blind friendly
library(RColorBrewer)
display.brewer.all(colorblindFriendly = T)

ggplot(data = movies, aes(x = genre, y = production_budget, fill = genre)) + 
  geom_violin(trim = F, alpha = 0.7) + 
  geom_point(colour = "white", size = 1, alpha = 0.1) + 
  stat_summary(fun.data="data_summary", col = "black") + 
  scale_fill_brewer(palette = "Dark2") + 
  thesis_theme + 
  ylab("Production budget (USD)") + 
  xlab("Genres") +
  theme(legend.position = "none")

#Viridis
library(viridisLite)

ggplot(data = movies, aes(x = production_budget, y = worldwide_gross)) +
  geom_hex() + 
  scale_x_log10() + 
  scale_y_log10() + 
  scale_fill_viridis() + 
  ylab("Production budget (USD)") + 
  xlab("Worldwide gross (USD)") +
  thesis_theme

#Wes Andersson
library(wesanderson)

movies %>% filter(distributor %in% top_6) %>% 
ggplot(aes(x = reorder(distributor, -production_budget, sum), y = production_budget, fill = distributor)) + 
  geom_col() + 
  ylab("Production budget (USD)") + 
  xlab("Distribution companies") + 
  scale_fill_manual(values = wes_palette("IsleofDogs1")) + 
  coord_flip() + 
  theme_bw() + 
  theme(legend.position = "none")

movies %>% filter(distributor %in% top_6) %>% 
ggplot(aes(x = reorder(distributor, -production_budget, sum), y = production_budget, fill = distributor)) + 
  geom_col() + 
  ylab("Production budget (USD)") + 
  xlab("Distribution companies") + 
  scale_fill_manual(values = wes_palette("IsleofDogs1")) + 
  coord_flip() + 
  theme_bw() + 
  theme(legend.position = "none")
```

### Arranging multiple plots
https://patchwork.data-imaginist.com/
```{r}
#Install package
install.packages('patchwork')
library(patchwork)

#Different layouts
A / D 

D / (A + C)

(A + B + C) / D

#Annotating
(A + B) / D + plot_annotation(tag_levels = 'A')
```


**Exercises**: Try to answer these questions below by making some relevant figures

1. Are there more cheap films?

 - Hint: create a histogram-plot with production_budget at the x-axis

```{r}
ggplot(movies, aes(x = production_budget)) +
  geom_histogram()
```
 
   
2. What genres make the highest return on investment?

 - Hints
   - calculate the ratio of worldwide gross profit and production budget
   - create a box-plot with x-axis representing the genre and y-axis representing the ratio
   - Rescale y-axis to log scale

```{r}
movies_ratio <- movies %>% mutate(ratio=worldwide_gross/production_budget)

ggplot(movies_ratio,aes(x = genre, y = ratio)) +
  geom_boxplot() +
   scale_y_log10()
```
   
3. Do expensive movies make more (gross) profit?

 - Hints
   - create a scatter plot with with the x-axis representing the production_budget and the y-axis representing the worldwide_gross profit
   - Rescale the x- and y-axis to log scale
   - add regression line with `stat_smooth()`

```{r}
# calculate the ratio of worldwide gross profit and production budget

movies_clean <- movies %>% filter(worldwide_gross!=0)

ggplot(movies_clean, aes(x = production_budget, y = worldwide_gross)) +
  geom_point(alpha=0.5) +
  stat_smooth(method = "lm", col = "red", se = FALSE)+
  scale_x_log10() + scale_y_log10()
```

## What to do when your big pipe thingo doesn't work?
```{r}
#load tidyverse
library(tidyverse)
#or make sure 'maggritr'
library(magrittr)
```


# Customisation: 

**Exercise**: Brain storm how to improve this plot

## scales and axes

We can uses the scales package for customising labels and ticks

```{r}
library(scales) #  

# calculate profit
movies_profit <- movies %>% 
   mutate(profit = worldwide_gross - production_budget)


ggplot(movies_profit, aes(x = production_budget, y = profit, color=mpaa_rating)) + # set category/color as mpaa_rating
   geom_point(alpha = .2) + # alpha: set transparency to .2 or 20%
   stat_smooth(method = "lm", col="red", se=FALSE) + # add linear regression model
    scale_x_log10(
   breaks = scales::trans_breaks("log10", function(x) 10^x),
   labels = scales::trans_format("log10", scales::math_format(10^.x))) + # log transform x axis and display number in math notation
   scale_y_continuous(labels = comma) + # log transform y axis and display number in natural notation
   theme_bw() + # set theme
   xlab("Production budget in US$") + # change x label
   ylab("Profit in US$") + # change y label
   theme(legend.position = "none") # remove the legend (mpaa_rating)
   
```


## facet_wrap

Add a single line to split by 

```{r}

ggplot(movies_profit, aes(x = production_budget, y = profit, color=mpaa_rating)) + # set category/color as mpaa_rating
   geom_point(alpha = .2) + # alpha: set transparency to .2 or 20%
   stat_smooth(method = "lm", col="red", se=FALSE) + # add linear regression model
    scale_x_log10(
   breaks = scales::trans_breaks("log10", function(x) 10^x),
   labels = scales::trans_format("log10", scales::math_format(10^.x))) + # log transform x axis and display number in math notation
   scale_y_continuous(labels = comma) + # log transform y axis and display number in natural notation
   theme_bw() + # set theme
   xlab("Production budget in US$") + # change x label
   ylab("Profit in US$") + # change y label
   theme(legend.position = "none") + # remove the legend (mpaa_rating)
   facet_wrap(~genre)  # split figures by genre
  
```


## plotly

The package plotly can be used to make plots interactive.

First save the plot as an object and add a fiEld for label:

```{r}

p <- ggplot(movies_profit, aes(x = production_budget, y = profit, color=mpaa_rating, label = movie)) + # set category/color as mpaa_rating
   geom_point(alpha = .2) + # alpha: set transparency to .2 or 20%
   facet_wrap(~genre) + # split figures by genre
    scale_x_log10(
   breaks = scales::trans_breaks("log10", function(x) 10^x),
   labels = scales::trans_format("log10", scales::math_format(10^.x))) + # log transform x axis and display number in math notation
   scale_y_continuous(labels = comma) + # log transform y axis and display number in natural notation
   theme_bw() + # set theme
   xlab("Production budget in US$") + # change x label
   ylab("Profit in US$") # change y label
```

It is then very easy to turn this into an interactive plot -- you'll need to click on "Viewer" rather than "Plots". You can hover over the points to see the deatils of each.

```{r}
library(plotly)
ggplotly(p)
```
