---
title: "Introduction to R"
author: "Instructors: Will Cornwell"
date: "27/1/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# What is R

- R consists of writing code
- Why write code?
  - reproducible - love not to your future self
- Code vs console
- Different clients -- Rstudio

#  The Rstudio environment 

Rstudio 

- into to panels

Writing code

- Code vs console vs Rmd
- Rmarkdown (code chunks, markdown)

# Packages


- Install package by using `install.packages("package-name")`
- Update packages by using `update.packages("package-name")`
- Load packages by using `library("package-name")`
- tidyverse is a useful collection of packages
- Possible to install packages from places other than cran (e.g. github, r-forge, local computer)
  - ways to install

Good instructions [here if needed](https://datacarpentry.org/R-ecology-lesson/#setup_instructions)

# Doing a full analysis yourself

```{r}
?read_csv

```


**Exercises:** Download course materials and knit demo. Go to the folder `lessons/10-R_intro/` and open the file `Intro_demo.Rmd` Make sure everyone can knit the demo. Take time to fix any issues that arise.


# Assigning objects

Code vs console vs Rmd

Use `<-` to assign something as an object in memory

- use `ls()` to see what's available in your environment 
- Rstudio environment panel
- Take errors seriously 

Running code vs assigning object

```{r}
12    # doesn't print anything
snakes <- 12
(snakes <- 12)  # but putting parenthesis around the call prints the value of `weight_kg`
snakes          # and so does typing the name of the object
```

<- vs = (outside vs inside function calls)


# Style

- the [tidyverse style guide](https://style.tidyverse.org/)
  - likes `_` between names
  - variable and function names should use only lowercase letters, numbers, and `_`. Use underscores (`_`) (so called snake case) to separate words within a name, e.g. `day_one`

**Exercises:**

1. Make a new variable containing the number 13
2. Make a new variable 1345 times the variable above (doing the calculation in code)
3. Make new variables containing the name of you and your partner 

**Resources:**

- ["Making new variables" on environmentalcomputing](http://environmentalcomputing.net/making-new-variables/)
- ["Creating R objects" on DataCarpentry](https://datacarpentry.org/R-ecology-lesson/01-intro-to-r.html)
- ["Software carpentry"'s intor to RStudio'](http://swcarpentry.github.io/r-novice-gapminder/01-rstudio-intro/index.html)

# Loading data

Where might you need to load data from (consider both sources and file types)? (Discuss with partner)

Tables are a special type

- tidy data: 1 observation per row, 1 variable per column

Comma Separated Files (csv) files are a common and desirable format. Why?

How do they work?

Use `read_csv` to load csv file

- intro to a function
- key arguments
- options for parsing 
- Loaded data need to be assigned as objects
- `read_csv` (in the tidyverse) is generally better than `read.csv`


```{r}
library(tidyverse)
movies <- read_csv("data/movie_profit/movie_profit.csv")
```

# Extract one column ($) 

 $ can also be used in assignment 

# Exercise 3: Add a column of worldwide net profit

 Subsetting by columns (select)

# Exercise 4: 
  
 1. Subset by columns to only movie title and worldwide net profit
 2. write this new dataframe to a file

# Logicals in R ( == < > | ! ) 

# characters in R ""

# Subset by rows (filter)

# Taking summary statistics: using function(...)

# Exercise 5:

 what's the mean of the worldwide gross for comedies?

#  Exercise 6: 
  
 what's the mean of the worldwide gross for R-rated comedies? 

#  Exercise 7: 

 what's the sd of the worldwide gross for Universal comedies? 

#  Exercise 8: 

 how many movies are in this database with a production budget of greater than $10,000,000? 

#  Exercise 9: 

 what percentage of movies in the database lost money (worldwide_gross - production_budget)? 



**Resources:**

You can likely load most data types into R

A few resources for loading other file types

- [readxl](https://readxl.tidyverse.org/reference/read_excel.html): Reading data direct from Excel
- [datapasta](https://cran.r-project.org/web/packages/datapasta/): when copying and pasting data from the web
- [googlesheets4](https://googlesheets4.tidyverse.org/): for reading data from google sheets



# Resources

Online courses:

- [Environmental Computing](http://environmentalcomputing.net/) 
- [Software carpentry's R novice lesson](http://swcarpentry.github.io/r-novice-gapminder/)
- [Datacarpentry's lesson on R for ecology](https://datacarpentry.org/R-ecology-lesson/)
- [A paper looking at how much R is used in Ecology](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.2567)
- [The tidyverse web page](https://www.tidyverse.org/)
- [Hadley Wickham's book R for data Science](https://r4ds.had.co.nz/)


Learning to code and need inspiration ?

- [Anyone can code](https://www.youtube.com/watch?v=qYZF6oIZtfc&list=PLzdnOPI1iJNe1WmdkMG-Ca8cLQpdEAL7Q)
- [Coding is the new literacy](https://www.youtube.com/watch?v=MwLXrN0Yguk&list=PLzdnOPI1iJNe1WmdkMG-Ca8cLQpdEAL7Q)
- [What most schools don't teach](https://www.youtube.com/watch?v=nKIu9yen5nc&feature=c4-overview-vl&list=PLzdnOPI1iJNe1WmdkMG-Ca8cLQpdEAL7Q)
- when coding saves you time, [from XKCD](http://xkcd.com/1205/)

Why good code matters

- [Why I want to write nice R code](http://nicercode.github.io/blog/2013-04-05-why-nice-code/)
- [Science has a credibility problem](http://www.economist.com/news/leaders/21588069-scientific-research-has-changed-world-now-it-needs-change-itself-how-science-goes-wrong)
