---
title: "Introduction to R"
author: "Instructors: Daniel Falster & Will Cornwell"
date: "17/12/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Welcome

(Dan, 9:00-9:30)

Logistics

- nametags: sticker and desk
- Sticky notes for help
- Work in pairs
  - help each other: Peer to peer is most efficient way of learning

Getting to know you:

- Whose used R before: <6 months, up to 2 years, more than 2 years
- Who has used a different language (e.g. Python)
- Discuss with partner: best & worst experience, goals, worries

Aims of this course:

- Introduction to R & Rstudio
- Getting data in and out of R 
- R objects and classes
- Packages
- Functions
- Organising your materials
- Data manipulation
- Data visualisation

Expectations

- We are practising scientists, also trained as instructors for software and data carpentry
  - here to guide your learning
  - it's *one* way to do it, not the only way
- We will show you the key ingredients for data analysis
  - we can't turn you into crack analysts in 2 days
  - start of a journey. Hopefully a fun one!
- We aim for boot camp to be inclusive
  - it's a diverse crowd, everyone has complimentary skills and experiences
  - there is no dumb question
  - anti-social behaviour not welcome

Learning to code and need inspiration ?

- [Anyone can code](https://www.youtube.com/watch?v=qYZF6oIZtfc&list=PLzdnOPI1iJNe1WmdkMG-Ca8cLQpdEAL7Q)
- [Coding is the new literacy](https://www.youtube.com/watch?v=MwLXrN0Yguk&list=PLzdnOPI1iJNe1WmdkMG-Ca8cLQpdEAL7Q)
- [What most schools don't teach](https://www.youtube.com/watch?v=nKIu9yen5nc&feature=c4-overview-vl&list=PLzdnOPI1iJNe1WmdkMG-Ca8cLQpdEAL7Q)
- when coding saves you time, [from XKCD](http://xkcd.com/1205/)

Why good code matters

- [Why I want to write nice R code](http://nicercode.github.io/blog/2013-04-05-why-nice-code/)
- [Science has a credibility problem](http://www.economist.com/news/leaders/21588069-scientific-research-has-changed-world-now-it-needs-change-itself-how-science-goes-wrong)

# A quick demo of some things you can do

(Dan, 9:30-10:00)

What is R

- R consists of writing code
- Why write code?
  - reproducible - love not to your future self
- Code vs console
- Different clients -- Rstudio

Demonstration of data analysis -- using the Sydney Beaches dataset

- see the file Intro_demo.Rmd

#  The Rstudio environment 

(Will, 10:00-10:30)

Rstudio 

- into to panels panels

Writing code

- Code vs console vs Rmd
- Rmarkdown (code chunks, markdown)

Packages

- installing
- tidyverse
- loading

**Exercises:** Install R (>3.5), Rstudio (1.1.463), and the `tidyverse` packages

Good instructions [here if needed](https://datacarpentry.org/R-ecology-lesson/#setup_instructions)

# Doing a full analysis yourself

(Will, 10:30-11:00)

**Exercises:** Download course materials and knit demo. Go to the folder `lessons/10-R_intro/` and open the file `Intro_demo.Rmd` Make sure everyone can knit the demo. Take time to fix any issues that arise.
  
# Morning tea

(11:00-11:30)

# Assigning objects

(Dan, 11:30-12:00)

Code vs console vs Rmd

Running code vs assigning object
Use <- to assign an object
Rstudio environment panel
errors

```{r}
12    # doesn't print anything
snakes <- 12
(snakes <- 12)  # but putting parenthesis around the call prints the value of `weight_kg`
snakes          # and so does typing the name of the object
```

<- vs = (outside vs inside function calls)
-> vs <- (-> works more naturally)

**Exercises:**

1. Make a new variable containing the number 13
2. Make a new variable 1345 times the variable above (doing the calculation in code)
3. Make new variables containing the name of you and your partner 

**Resources:**

- ["Making new variables" on environmentalcomputing](http://environmentalcomputing.net/making-new-variables/)
- ["Creating R objects" on DataCarpentry](https://datacarpentry.org/R-ecology-lesson/01-intro-to-r.html)

# Loading data

(Dan, 12:00-12:30)

Loaded data as objects

Where might you need to load data from (consider both sources and file types)? (Discuss with partner)

Tables are a special type

- tidy data: 1 observation per row, 1 variable per column

Comma Separated Files (csv) files are a common and desirable format. Why?

How do they work?

Use `read_csv` to load csv file

- intro to a function
- key arguments
- options for parsing 

**Exercises:**

1. Load the file `10-R_intro/data/temperature.csv` into R and save as a variable
2. Create a csv file from either Excel or Google Sheets
3. Open your csv file in the Rstudio editor (or another text editor)


**Resources:**

Resources for loading other file types

- [readxl](https://readxl.tidyverse.org/reference/read_excel.html): Reading data direct from Excel
- [datapasta](https://cran.r-project.org/web/packages/datapasta/): when copying and pasting data from the web
- [googlesheets4](https://googlesheets4.tidyverse.org/): for reading data from google sheets


# Lunch 

(12:30 - 13:30)

# Object types I

(Will, 13:30-14:30)

- data frame
   - columns
   - $
- vector
   - []
   - :

**Exercises:**

- data types
  - character
  - integer
  - numeric
  - boolean
  - dates
- variables
- matrix
- lists
  - non-table structures
  - calling via names, numbers

# Exploring data: useful vocabulary

(Dan, 14:30-15:00)


- vector
  - numeric: mean, max, length
  - character: unique

**Exercises:**

1. load data, 
  - take column, first 5 values, take mean
  - take all column x by value column y, first 5 values, take mean
  - multiply columns together
  - multiply columns together then sum
2. (for some data) Calculate sd and check against sd function

# Afternoon tea

(15:00 - 15:30)

# Exploring data

**Exercises:**

# More complicated functions and how to use them 

(Dan. 16:00-16:30)

- arguments for functions
- default values
- = 
- outputs of functions
- lm
- formula

**Exercises:**

# Optional extras

(if time permits)

Where to get help

- google
- stack overflow
- reproducible examples


# Resources

- [Environmental Computing](http://environmentalcomputing.net/) 
- [Datacarpentry's lesson on R for ecology](https://datacarpentry.org/R-ecology-lesson/)
- [A paper looking at how much R is used in Ecology](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.2567)
- [The tidyverse web page](https://www.tidyverse.org/)
- [Hadley Wickham's book R for data Science](https://r4ds.had.co.nz/)
