---
title: "Introduction to R"
author: "Instructors: Daniel Falster & Will Cornwell"
date: "17/12/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Welcome

(Dan, 9:00-9:30)

Logistics

- nametags: sticker and desk
- Sticky notes for help
- Work in pairs
  - help each other: Peer to peer is most efficient way of learning

Getting to know you:

- Whose used R before: <6 months, up to 2 years, more than 2 years
- Who has used a different language (e.g. Python)
- Discuss with partner: best & worst experience, goals, worries

Aims of this course:

- Introduction to R & Rstudio
- Getting data in and out of R 
- R objects and data types
- Packages
- Functions
- Organising your materials in Projects
- Data manipulation
- Data visualisation

Expectations

- We are practising scientists, also trained as instructors for software and data carpentry
  - here to guide your learning
  - it's *one* way to do it, not the only way
- We will show you the key ingredients for data analysis
  - we can't turn you into crack analysts in 2 days
  - start of a journey. Hopefully a fun one!
- We aim for boot camp to be inclusive
  - it's a diverse crowd, everyone has complimentary skills and experiences
  - there is no dumb question
  - anti-social behaviour not welcome
- Live coding
  - We're going to do a lot of live coding -> why?
  - Materials will be circulated, but we don't have fully worked files for you. There are lots of these already available (see Resources sections), including our own website: [Environmental Computing](http://environmentalcomputing.net/) 
  - Use the circulated files to take notes as you follow along
  - We'll circulate our files after 

# A quick demo of some things you can do

(Dan, 9:30-10:00)

What is R

- R consists of writing code
- Why write code?
  - reproducible - love not to your future self
- Code vs console
- Different clients -- Rstudio

Demonstration of data analysis -- using the Sydney Beaches dataset

- see the file Intro_demo.Rmd

#  The Rstudio environment 

(Will, 10:00-10:30)

Rstudio 

- into to panels panels

Writing code

- Code vs console vs Rmd
- Rmarkdown (code chunks, markdown)

Packages


- Install package by using `install.packages("package-name")`
- Update packages by using `update.packages("package-name")`
- Load packages by using `library("package-name")`
- tidyverse is a useful collection of packages
- Possible to install packages from places other than cran (e.g. github, r-forge, local computer)
  - ways to install

**Exercises:** Install R (>3.5), Rstudio (1.1.463), and the `tidyverse` packages

Good instructions [here if needed](https://datacarpentry.org/R-ecology-lesson/#setup_instructions)

# Doing a full analysis yourself

```{r}
?read_csv

```
(Will, 10:30-11:00)

**Exercises:** Download course materials and knit demo. Go to the folder `lessons/10-R_intro/` and open the file `Intro_demo.Rmd` Make sure everyone can knit the demo. Take time to fix any issues that arise.
  
# Morning tea

(11:00-11:30)

# Assigning objects

(Dan, 11:30-12:00)

Code vs console vs Rmd

Use `<-` to assign something as an object in memory

- use `ls()` to see what's available in your environment 
- Rstudio environment panel
- Take errors seriously 

Running code vs assigning object

```{r}
12    # doesn't print anything
snakes <- 12
(snakes <- 12)  # but putting parenthesis around the call prints the value of `weight_kg`
snakes          # and so does typing the name of the object
```

<- vs = (outside vs inside function calls)
-> vs <- (-> works more naturally)

use `rm()` to delete objects

Style

- the [tidyverse style guide](https://style.tidyverse.org/)
  - likes `_` between names
  - variable and function names should use only lowercase letters, numbers, and `_`. Use underscores (`_`) (so called snake case) to separate words within a name, e.g. `day_one`

**Exercises:**

1. Make a new variable containing the number 13
2. Make a new variable 1345 times the variable above (doing the calculation in code)
3. Make new variables containing the name of you and your partner 

**Resources:**

- ["Making new variables" on environmentalcomputing](http://environmentalcomputing.net/making-new-variables/)
- ["Creating R objects" on DataCarpentry](https://datacarpentry.org/R-ecology-lesson/01-intro-to-r.html)
- ["Software carpentry"'s intor to RStudio'](http://swcarpentry.github.io/r-novice-gapminder/01-rstudio-intro/index.html)

# Loading data

(Dan, 12:00-12:30)

Where might you need to load data from (consider both sources and file types)? (Discuss with partner)

Tables are a special type

- tidy data: 1 observation per row, 1 variable per column

Comma Separated Files (csv) files are a common and desirable format. Why?

How do they work?

Use `read_csv` to load csv file

- intro to a function
- key arguments
- options for parsing 
- Loaded data need to be assigned as objects
- `read_csv` (in the tidyverse) is generally better than `read.csv`

**Exercises:**

1. Load the file `10-R_intro/data/sydneybeaches.csv` into R and save as a variable called `data_beaches`
2. Create a csv file from either Excel or Google Sheets
3. Open your csv file in the Rstudio editor (or another text editor)

```{r}
x <- TRUE
class(x)
```



**Resources:**

You can likely load most data types into R

A few resources for loading other file types

- [readxl](https://readxl.tidyverse.org/reference/read_excel.html): Reading data direct from Excel
- [datapasta](https://cran.r-project.org/web/packages/datapasta/): when copying and pasting data from the web
- [googlesheets4](https://googlesheets4.tidyverse.org/): for reading data from google sheets

# Lunch 

(12:30 - 13:30)

# Object types I

(Will, 13:30-14:30)

- data frame
   - columns
   - $
- vector
   - []
   - :
- data types
  - character
  - integer
  - numeric
  - boolean
  - dates
- variables
- matrix
- lists
  - non-table structures
  - calling via names, numbers




**Exercises:**

create a new column in the data_beaches data frame and change the units of Enterococci from cfu_per100ml to Enterococci_cfu_per_10ml

the Enterococci_cfu_per_100ml variable to Enterococci_cfu_per_10ml

```{r}
data_beaches <- read_csv("data/sydneybeaches.csv")
dim(data_beaches)

data_beaches[,1:3]
sort(unique(data_beaches$Site),decreasing = TRUE)
```
# Exploring data: useful vocabulary

(Dan, 14:30-15:00)

We're now going to check out a bunch of really useful R functions and see how they work, to explore some data in the "Sydney Beaches" data set. 

First load the data from `data/sydneybeaches.csv` and save as `data_beaches`


To look at columns of numeric data, these functions can be helpful:

- min, max, range, sum, length, mean, sd

To look at columns of categorical data, these functions can be helpful:

- unique, sort, length

These functions can be used on whole dataframes

- nrow
- dim
- colSum, rowSum

**Exercises:**

1. Load the beaches data, then 
  - extract the column `Enterococci_cfu_per_100ml` and save as a new variable
  - find the min and max count
  - find the mean
  - find the mean of the first 100 records
2. Using the same data, extract the column `Site` and save as a new variable
  - find all unique values
  - find all unique values and sort them
3. using the whole dataframe
  - How many rows and columns does it have
  - Make a new dataset using only the first 10 rows
  - Make a new dataset using only the first 3 columns

# Afternoon tea

(15:00 - 15:30)

# More complicated functions and how to use them 

(Dan. 15:30-16:00)

Investigate function structure, using the function `lm` as an example

- function structure
  - name
  - body
  - arguments
- investigating structure of function f 
  - `args(f)`
  - `body(f)` or just `f`
  - return value
- ?help
- named vs ordered arguments
- default arguments

```{r}
model_fit<-lm(iris$Sepal.Length~iris$Petal.Length)
```



# Q & A 

(16:00 - 16:30)

**Exercises:**

# Optional extras

(if time permits)

Where to get help

- google
- stack overflow
- reproducible examples


# Resources

Online courses:

- [Environmental Computing](http://environmentalcomputing.net/) 
- [Software carpentry's R novice lesson](http://swcarpentry.github.io/r-novice-gapminder/)
- [Datacarpentry's lesson on R for ecology](https://datacarpentry.org/R-ecology-lesson/)
- [A paper looking at how much R is used in Ecology](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.2567)
- [The tidyverse web page](https://www.tidyverse.org/)
- [Hadley Wickham's book R for data Science](https://r4ds.had.co.nz/)


Learning to code and need inspiration ?

- [Anyone can code](https://www.youtube.com/watch?v=qYZF6oIZtfc&list=PLzdnOPI1iJNe1WmdkMG-Ca8cLQpdEAL7Q)
- [Coding is the new literacy](https://www.youtube.com/watch?v=MwLXrN0Yguk&list=PLzdnOPI1iJNe1WmdkMG-Ca8cLQpdEAL7Q)
- [What most schools don't teach](https://www.youtube.com/watch?v=nKIu9yen5nc&feature=c4-overview-vl&list=PLzdnOPI1iJNe1WmdkMG-Ca8cLQpdEAL7Q)
- when coding saves you time, [from XKCD](http://xkcd.com/1205/)

Why good code matters

- [Why I want to write nice R code](http://nicercode.github.io/blog/2013-04-05-why-nice-code/)
- [Science has a credibility problem](http://www.economist.com/news/leaders/21588069-scientific-research-has-changed-world-now-it-needs-change-itself-how-science-goes-wrong)
