---
title: "Data Manipulation and Visualization (2)"
author: "Instructors: Daniel Falster, Will Cornwell, Dony Indiarto"
date: "Feb 2020"
output: html_document
---


# Introduction to gapminder data (Dan)

For day 2's exercises, we're going to use the famous gapminder dataset, maintained at [gapminder.org](http://gapminder.org)

This dataset tells a grand story about human development across countries, and was made famous in TED talks by Hans Rosling -- eg. see https://www.ted.com/talks/hans_rosling_the_best_stats_you_ve_ever_seen#t-126589 (start 2:10)

**Exercise**: With your partner, review the `README` for the data at `data/gapminder/readme.md` to see the variables included. 

# Tidy Data concept (Dony)

- What is tidy data?
- Converting from long to wide data
- see slides in `slides/tidy_data_and_pivot_tables.pdf`

Read and inspect the structure of `life_expectancy_years.csv`

```{r}
library(tidyverse)
le <- read_csv("data/gapminder/life_expectancy_years.csv") # read the worldwide life expectancy data

#skimr::skim(le) #summarise using the `skimr

#View(le) #An overview of the data
```


Make the 'life expectancy' data tidy with `dplyr::pivot_longer( )`

```{r}
le_long<- 
  pivot_longer(data = le, # the data
               
                      # pivot all column into longer format except 'country'
                      cols=-country, 
               
                      #the name of the column made from the data stored in the col.names
                      names_to="year", 
                  
                      #the name of the column made from the data stored in the cells
                      values_to = "life_exp") 

View(le_long)
```

Make a figure of life expectancy trend over time

```{r}
 ggplot(le_long,                                   #data
 aes(x = year, y = life_exp, group = country)) +   #aesthetics
 geom_line() +	                                   #type of plot
 theme(legend.position ="none")                    #remove legend

```

Make the long 'life expectancy' data back to its original form with `dplyr::pivot_wider( )`

```{r}
le_wide <- pivot_wider(data = le_long,           #the data
                        names_from = "year",     #column to get the column names from
                        values_from = "life_exp")#column to get the values from

View(le_wide) 
```

**Exercises**:

Make these data tidy:

- `data\gapminder\children_per_woman_total_fertility.csv`
- `data\gapminder\income_per_person_gdppercapita_ppp_inflation_adjusted.csv`


```{r}

data_fertility <- 
  read_csv("data/gapminder/children_per_woman_total_fertility.csv") %>%   
  pivot_longer(-country, names_to = "year", values_to = "fertility")

data_gdp <- read_csv("data/gapminder/income_per_person_gdppercapita_ppp_inflation_adjusted.csv") %>%   
  pivot_longer(-country, names_to = "year", values_to = "gdp_per_capita")

```


# Advanced data manipulation with `dplyr` (Dan)

We're now going to focus in on some more advanced features of dplyr. First, we need to load a series of tables and use `pivot_longer` to reorganise them. 

```{r}
# Fertility
data_fertility <- read_csv("data/gapminder/children_per_woman_total_fertility.csv") %>%   
  pivot_longer(-country, names_to = "year", values_to = "fertility")
  
# Population
data_population <- read_csv("data/gapminder/population_total.csv") %>%
  pivot_longer(-country, names_to = "year", values_to = "population")

# gdp per capita
data_gdp <- read_csv("data/gapminder/income_per_person_gdppercapita_ppp_inflation_adjusted.csv") %>%   
  pivot_longer(-country, names_to = "year", values_to = "gdp_per_capita")

# life expectancy
data_life_exp <- read_csv("data/gapminder/life_expectancy_years.csv") %>%
  pivot_longer(-country, names_to = "year", values_to = "life_expectancy")

# continents
data_continent <-  read_csv("data/gapminder/continents.csv")
  
```

But wouldn't it be nice to merge these into a single dataset?

That's where `dplyr`s collection of `join` functions come in  

## Joining tables

https://dplyr.tidyverse.org/reference/join.html

There's three broad types of joins

1. Mutating joins combine variables from the two data.frames
2. Filtering joins keep cases from the left-hand data.frame
3. Nesting joins create a list column of data.frames

We're going to focus here on one type of Mutating join:

- `left_join()`: return all rows from x, and all columns from x and y. Rows in x with no match in y will have NA values in the new columns. If there are multiple matches between x and y, all combinations of the matches are returned.

The key to making a join work is to identify the variable from each table that link the tables together. 

Let's say we want add information about each countries continent into `data_population`

```{r}
left_join(data_population, data_continent, by="country")
```

Note that

- the order of tables matters
- it still works without the "by" argument, because it guesses which columns are in common
- you don't need to have the same named columns but this makes life much easier.

We can also join by two variables 

```{r}
left_join(data_population, data_fertility)
```

or

```{r}
left_join(data_population, data_fertility, by = c("country", "year"))
```

**Exercise:** 

1. Join the two data frames `data_population` and `data_life_exp`

```{r}
left_join(data_population, data_life_exp)

```

2. Join the three data frames `data_population`, `data_life_exp` and `data_continent`. 
  - hint: the join functions can only handle one table at a time
  - Optional: reorder so that continent is the first variable

```{r}
data_population %>% 
  left_join(data_life_exp) %>%
  left_join( data_continent) %>% 
  select(continent, country, year, everything())
```


----

Now using join and pipes, we can built a big table of data containing all variables

```{r}
data <- 
  data_fertility %>% 
  left_join( data_population) %>% 
  left_join( data_gdp ) %>% 
  left_join(data_life_exp) %>%
  left_join(data_continent) %>% 
  # reorganise data
  select(continent, country, year, everything()) %>% 
  arrange(continent, country, year) %>% 
  mutate(year = as.integer(year))

```

I'd probably read in the csv tables and join them straight in, without saving as intermediate objects

```{r}

data <- 
  read_csv("data/gapminder/children_per_woman_total_fertility.csv") %>%   
  pivot_longer(-country, names_to = "year", values_to = "fertility") %>%
  # Add in population
  left_join(
    read_csv("data/gapminder/population_total.csv") %>%
              pivot_longer(-country, names_to = "year", values_to = "population")
  ) %>%
  # Add in gdp per capita
  left_join(
    read_csv("data/gapminder/income_per_person_gdppercapita_ppp_inflation_adjusted.csv") %>%   
              pivot_longer(-country, names_to = "year", values_to = "gdp_per_capita")
  ) %>%
  # Add in life expectancy
  left_join(
    read_csv("data/gapminder/life_expectancy_years.csv") %>%
              pivot_longer(-country, names_to = "year", values_to = "life_expectancy")
  ) %>%
  # Add in life continent
  left_join(
    read_csv("data/gapminder/continents.csv")
  ) %>% 
  # reorganise data
  select(continent, country, year, everything()) %>% 
  arrange(continent, country, year) %>% 
  mutate(year = as.integer(year))
```

Now we're going to do some analysis on this dataset using `dplyr`. 

## Refresher

First let's refresh some of the commands we learnt yesterday: `select`, `filter`, `rename`, `arrange`, `mutate`. 

**Exercises:**

1. Your collaborator asks for a copy of the dataset with only data from Europe since 1900, with variables `country`, `year`, `fertility`, `population`. Use pipes and dplyr commands to create a csv to send them. 
  - Hint: the command `write_csv` will save a table and works with the pipe.
  
  
```{r}
data %>% 
  filter(year > 1900, continent == "Europe") %>% 
  select(country, year, fertility, population) %>% 
  write_csv("output/europe.csv")

```


2. The journal where you are submitting your paper wants a table summarising the dataset, giving the 
  - number of countries & continents
  - time frame
  - mean of variables fertility and life_expectancy
  - hints: the functions summarise and n_distinct will be useful

```{r}
data %>%
  summarise(
    start = min(year),
    end = max(year),
    countries  = n_distinct(country),
    fertility = mean(fertility),
    life_expectancy = mean(life_expectancy)
  )
```


Your answer should look like this:

```
  start   end countries fertility life_expectancy
  1800   2018       184      5.40            43.0
```

3. Try the same as above, but with answers for each continent including in the data
  - you won't be able to solve this, without material from the next section. But think about what you would in principle like to do.


```{r}

```

## Group_by 

When you want to run an operation on subgroups in the data, we turn to `dplyr's` `group_by` command. This turbo-charges other commands like `summarise` & `mutate`. 

Let's revisit the challenge above: The journal where you are submitting your paper wants a table summarising the dataset, giving **for each continent** the 
  - number of countries
  - time frame
  - mean of variables fertility and life_expectancy
  - hints: the functions summarise and n_distinct will be useful


```{r}
data %>%
  group_by(continent) %>% 
  summarise(
    start = min(year),
    end = max(year),
    countries  = n_distinct(country),
    fertility = mean(fertility),
    life_expectancy = mean(life_expectancy)
  )
```

Group_by can also be used to calculate mutations at the group level. Let's image you wanted to calculate each countries population as a percentage of its current population. 

```{r}
data2 <- 
  data %>%
  group_by(country) %>% 
  mutate(
    population_percent = population / last(population) 
  ) %>%
  ungroup() %>%
  select(-fertility, -gdp_per_capita, -life_expectancy )
```


Note, when creating a new dataset it is also important to remove the grouping variable using `ungroup`



**Exercise**: Use the commands `group_by`  command to

1. Calculate average values of life_expectancy and population for each country in the period 1990-1999
  - hint: you'll also need `filter` and `summarise`

```{r}
data %>%
  filter(year >= 1990, year < 2000) %>%
  group_by(country) %>%
  summarise(
    population = mean(population),
    life_expectancy = mean(life_expectancy)
  )
```

Your answer should took like this:

```
# A tibble: 184 x 3
   country             population life_expectancy
   <chr>                    <dbl>           <dbl>
 1 Afghanistan           16210000            51.2
 2 Albania                3164000            73.9
 3 Algeria               28520000            72.3
 ...
```



2. Calculate average values of life_expectancy and population for each country in the period 1990-1999
  - hint: you'll also need `filter` and `summarise`

```{r}
data %>%
  filter(year >= 1990, year < 2000) %>%
  group_by(country) %>%
  summarise(
    population = mean(population),
    life_expectancy = mean(life_expectancy)
  )
```

Your answer should took like this:

```
# A tibble: 184 x 3
   country             population life_expectancy
   <chr>                    <dbl>           <dbl>
 1 Afghanistan           16210000            51.2
 2 Albania                3164000            73.9
 3 Algeria               28520000            72.3
 ...
```


## Handling NA's

Now we'll briefly look at what to do when you need to 


First, let's identify where we have missing data. The function complete.cases is handy:

```{r}
complete.cases(data)
```

We can then use filter to select the rows


```{r}
i <- complete.cases(data)
data %>% filter(!i)
```

Looks liek some NA's in continent. Let's find which countries
```{r}
data %>% 
  filter(is.na(continent)) %>% 
  select(country) %>% 
  distinct() %>%
  pull(country)
```

Now let's fix some of this

```{r}
data2 <- 
  data %>%
  mutate(continent = ifelse(country == "Russia", "Europe", continent),
         continent = ifelse(is.na(continent), "Islands", continent)
  )
```

Instead of an ifelse, we could also use the `replace_na` function

```{r}

data2 <- 
  data %>%
  mutate(continent = ifelse(country == "Russia", "Europe", continent)) %>%
  replace_na(list(continent="Islands"))
```


# Advanced data visualisation with `ggplot2` (Will)

This session will show you some more advanced plotting techniques, building on the basic plots from Day 1 and using the gapminder data


## Colour schemes

### Continuous colour and discrete colour

### Colour-blindness

https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html

```{r}
library(viridis)
library(ggplot2)
ggplot(data.frame(x = rnorm(10000), y = rnorm(10000)), aes(x = x, y = y)) +
  geom_hex() + coord_fixed() +
  scale_fill_viridis(option="D") + theme_bw()
```



```{r}
p <- ggplot(mtcars, aes(wt, mpg))
p + geom_point(size=4, aes(colour = factor(cyl))) +
    scale_color_viridis(discrete=TRUE) +
    theme_bw()

```


```{r}
library(ggsci)
ggplot(subset(diamonds, carat >= 2.2),
       aes(x = table, y = price, colour = cut)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", alpha = 0.1, size = 1, span = 1,se=FALSE) +
  theme_bw() + scale_color_npg()


```
```{r}

ggplot(subset(diamonds, carat >= 2.2),
       aes(x = table, y = price, colour = cut)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", alpha = 0.1, size = 1, span = 1,se=FALSE) +
  theme_bw() + scale_color_aaas()

```

## one other thing about font size

```{r}
ggplot(subset(diamonds, carat >= 2.2),
       aes(x = table, y = price, colour = cut)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", alpha = 0.1, size = 1, span = 1,se=FALSE) +
  theme_bw() + scale_color_aaas()+
  theme(text = element_text(size=18)) 
```



## Practice dplyr and ggplot 1

1. Filter to only data from the year 1900

2. make a scatter plot with 1) gdp_per_capita as the x-axis 2) life_expectancy as the y-axis 3) continent as the colour of the points, 4) population as the size of the points

3. modify the axes to be be more informative

4. use the theme you think looks best

5. make the font size of the axes 18 (for a presentation)

```{r}
a<-data %>% 
  filter(year==1900) %>%
  ggplot(aes(gdp_per_capita, life_expectancy, colour = continent, size=population,label=country)) + 
  geom_point() + 
  scale_x_log10() + scale_y_log10() +
  labs(x = "GDP per capita ($)", y = "life expectancy (yr)") + theme_classic() +
  theme(text = element_text(size=18)) 
```

## One more logical

```{r}
 "a" %in% c("b","c","e","g")
```
```{r}
 "a" %in% c("b","c","e","g","a")
```



## Facets

1. Filter to data from three years:  1800, 1850,1900, 1950,2000

2. Remove countries that have NA for continent

3. make a scatter plot with 1) year as the x-axis 2) life expectancy as the y-axis 3) log10(population) as the colour of the points with viridis as the colour scheme 

4. facet by continent

```{r}
data %>% 
  filter(year %in% c(1800,1850,1900,1950,2000)) %>%
  filter(!is.na(continent)) %>%
  ggplot(aes(year, life_expectancy, colour = log10(population))) + 
  theme_classic()+
  geom_point(alpha = 0.7) +
  scale_x_log10() + 
  scale_color_viridis()+
  labs(x = "GDP per capita ($)", 
       y = "life expectancy (yr)",
       caption = "(based on data from gapminder)",
       color= "Continent (log)") + 
  theme_classic() +
  facet_wrap(~continent)
```


## Custom Themes


Can add a theme onto plot

```{r}
mynamestheme <- theme( 
                  axis.title = element_text(family = "Helvetica", size = (15), colour = "steelblue4"),
                  axis.text = element_text(family = "Courier", colour = "cornflowerblue", size = (15)),
                  legend.text = element_text(family = "Helvetica", size = (15), colour = "pink"),
                  legend.title = element_text(family = "Helvetica", size = (15), colour = "green",face="italic"),
                  legend.position="left")
  
a + mynamestheme 
```

### Does the correlation between gdp and fertility get stronger or weaker through time?


1. filter year to only values for 1800,1850, 1900, 1950, and 2000

2. filter out values without a continent entry

3. make a scatter plot with x axis = gdp_per_capita and y axis = fertility, colour based on continent (log scale the x-axis, use a discrete color scale)

4. facet by year

```{r}
head(data)

data %>% 
  filter(year %in% c(1800,1850,1900,1950,2000)) %>%
    filter(!is.na(continent)) %>%
  ggplot(aes(gdp_per_capita, fertility, colour = continent)) + 
  theme_classic()+
  geom_point(alpha = 0.7) +
  scale_x_log10() + 
  scale_color_viridis(discrete = TRUE) + 
  theme_classic() +
  facet_wrap(~year)
```

### Animations with gganimate

https://github.com/thomasp85/gganimate

```{r}
library(gganimate)

country_colors <- readRDS("data/gapminder/country_colours.rds")

data %>% 
  filter(!is.na(continent)) %>% 
  ggplot(aes(gdp_per_capita, life_expectancy, size = population, colour = country)) +
  theme_classic() +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  ease_aes('linear')
```


# Reproducible research (Dan)

**Exercises:** Discuss with your partner 

- what is meant by Reproducible research?
- who benefits from reproducible research?
- what can you do do make your research reproducible?

