---
title: "Data Manipulation and Visualization (2)"
author: "Instructors: Daniel Falster, Will Cornwell, Dony Indiarto"
date: "Feb 2020"
output: html_document
---


# Introduction to gapminder data (Dan)

For day 2's exercises, we're going to use the famous gapminder dataset, maintained at [gapminder.org](http://gapminder.org)

**Exercise**: With your partner, review the `README` for the data at `data/gapminder/readme.md` to see the variables included. 


# Tidy Data concept (Dony)

- What is tidy data?
- Converting from long to wide data
- see slides in `slides/tidy_data_and_pivot_tables.pdf`

Read and inspect the structure of `life_expectancy_years.csv`

```{r}
library(tidyverse)
le <- read_csv("data/gapminder/life_expectancy_years.csv") # read the worldwide life expectancy data

#skimr::skim(le) #summarise using the `skimr

#View(le) #An overview of the data
```


Make the 'life expectancy' data tidy with `dplyr::pivot_longer( )`

```{r}
le_long<- 
  pivot_longer(data = le, # the data
               
                      # pivot all column into longer format except 'country'
                      cols=-country, 
               
                      #the name of the column made from the data stored in the col.names
                      names_to="year", 
                  
                      #the name of the column made from the data stored in the cells
                      values_to = "life_exp") 

View(le_long)
```

Make a figure of life expectancy trend over time

```{r}
 ggplot(le_long,                                   #data
 aes(x = year, y = life_exp, group = country)) +   #aesthetics
 geom_line() +	                                   #type of plot
 theme(legend.position ="none")                    #remove legend

```

Make the long 'life expectancy' data back to its original form with `dplyr::pivot_wider( )`

```{r}
le_wide <- pivot_wider(data = le_long,           #the data
                        names_from = "year",     #column to get the column names from
                        values_from = "life_exp")#column to get the values from

View(le_wide) 
```

**Exercises**:

Make these data tidy:

- `data\gapminder\children_per_woman_total_fertility.csv`
- `data\gapminder\income_per_person_gdppercapita_ppp_inflation_adjusted.csv`


# Advanced data manipulation with `dplyr` (Dan)

We're now going to focus in on some more advanced features of dplyr. First, we need to load a series of tables and use `pivot_longer` to reorgaise them. 

```{r}
# Fertility
data_fertility <- read_csv("data/gapminder/children_per_woman_total_fertility.csv") %>%   
  pivot_longer(-country, names_to = "year", values_to = "fertility")
  
# Population
data_population <- read_csv("data/gapminder/population_total.csv") %>%
  pivot_longer(-country, names_to = "year", values_to = "population")

# gdp per capita
data_gdp <- read_csv("data/gapminder/income_per_person_gdppercapita_ppp_inflation_adjusted.csv") %>%   
  pivot_longer(-country, names_to = "year", values_to = "gdp_per_capita")

# life expectancy
data_life_exp <- read_csv("data/gapminder/life_expectancy_years.csv") %>%
  pivot_longer(-country, names_to = "year", values_to = "life_expectancy")

# continents
data_continent <-  read_csv("data/gapminder/continents.csv")
  
```

But wouldn't it be nice to merge these into a single dataset?

That's where `dplyr`s collection of `join` functions come in  

## Joining tables

https://dplyr.tidyverse.org/reference/join.html

There's three broad types of joins

1. Mutating joins combine variables from the two data.frames
2. Filtering joins keep cases from the left-hand data.frame
3. Nesting joins create a list column of data.frames

We're going to focus here on one type of Mutating join:

- `left_join()`: return all rows from x, and all columns from x and y. Rows in x with no match in y will have NA values in the new columns. If there are multiple matches between x and y, all combinations of the matches are returned.

The key to making a join work is to indetify the variable from each table that link the tables together. 

Let's say we want add infomration about each countried continent into `data_population`

```{r}
left_join(data_population, data_continent, by="country")
```

Note that

- the order of tables matters
- it still works without the "by" argument, because it guesses which columns are in common
- you don't need to have the same named columns but this makes life much easier.

We can also join by two variables 

```{r}
left_join(data_population, data_fertility)
```

or

```{r}
left_join(data_population, data_fertility, by = c("country", "year"))
```

**Exercise:** 

1. Join the two data frames `data_population` and `data_life_exp`
2. Join the three data frames `data_population`, `data_life_exp` and `data_continent`


----

Now using join and pipes, we can built a big table of data containing all variables

```{r}

data <- 
  read_csv("data/gapminder/children_per_woman_total_fertility.csv") %>%   
  pivot_longer(-country, names_to = "year", values_to = "fertility") %>%
  # Add in population
  left_join(
    read_csv("data/gapminder/population_total.csv") %>%
              pivot_longer(-country, names_to = "year", values_to = "population")
  ) %>%
  # Add in gdp per capita
  left_join(
    read_csv("data/gapminder/income_per_person_gdppercapita_ppp_inflation_adjusted.csv") %>%   
              pivot_longer(-country, names_to = "year", values_to = "gdp_per_capita")
  ) %>%
  # Add in life expectancy
  left_join(
    read_csv("data/gapminder/life_expectancy_years.csv") %>%
              pivot_longer(-country, names_to = "year", values_to = "life_expectancy")
  ) %>%
  # Add in life continent
  left_join(
    read_csv("data/gapminder/continents.csv")
  ) %>% 
  # reorganise data
  select(continent, country, year, everything()) %>% 
  arrange(continent, country, year) %>% 
  mutate(year = as.integer(year))
```

What about Na's?

```{r}
filter(data, is.na(continent))
filter(data, is.na(continent)) %>% pull(country) %>% unique()

```

## Group_by (summarise, mutate)


## What to do when your big pipe thingo doesn't work?


# Advanced data visualisation with `ggplot2` (Will)

This session will show you some more advanced plotting techniques, building on the basic plots from Day 1 and using the gapminder data


## Color schemes

### Continuous color and discrete color

### Colorblindness

https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html

```{r}
library(viridis)
library(ggplot2)
ggplot(data.frame(x = rnorm(10000), y = rnorm(10000)), aes(x = x, y = y)) +
  geom_hex() + coord_fixed() +
  scale_fill_viridis(option="D") + theme_bw()
```



```{r}
p <- ggplot(mtcars, aes(wt, mpg))
p + geom_point(size=4, aes(colour = factor(cyl))) +
    scale_color_viridis(discrete=TRUE) +
    theme_bw()

```


```{r}
library(ggsci)
ggplot(subset(diamonds, carat >= 2.2),
       aes(x = table, y = price, colour = cut)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", alpha = 0.1, size = 1, span = 1,se=FALSE) +
  theme_bw() + scale_color_npg()


```
```{r}

ggplot(subset(diamonds, carat >= 2.2),
       aes(x = table, y = price, colour = cut)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", alpha = 0.1, size = 1, span = 1,se=FALSE) +
  theme_bw() + scale_color_aaas()

```

## one other thing about font size

```{r}
ggplot(subset(diamonds, carat >= 2.2),
       aes(x = table, y = price, colour = cut)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", alpha = 0.1, size = 1, span = 1,se=FALSE) +
  theme_bw() + scale_color_aaas()+
  theme(text = element_text(size=18)) 
```



## Practice dplyr and ggplot 1

1. Filter to only data from the year 1900

2. make a scatter plot with 1) gdp_per_capita as the x-axis 2) life_expectancy as the y-axis 3) continent as the colour of the points, 4) population as the size of the points

3. modify the axes to be be more informative

4. use the theme you think looks best

5. make the font size of the axes 18 (for a presentation)

```{r}

```

## One more logical

```{r}
 "a" %in% c("b","c","e","g")
```
```{r}
 "a" %in% c("b","c","e","g","a")
```



## Facets

1. Filter to data from three years:  1800, 1850,1900, 1950,2000

2. Remove countries that have NA for continent

3. make a scatter plot with 1) year as the x-axis 2) life expectancy as the y-axis 3) log10(population) as the colour of the points with viridis as the color scheme 

4. facet by continent

```{r}

```


## Custom Themes


Can add a theme onto plot

```{r}
mynamestheme <- theme( 
                  axis.title = element_text(family = "Helvetica", size = (15), colour = "steelblue4"),
                  axis.text = element_text(family = "Courier", colour = "cornflowerblue", size = (15)),
                  legend.text = element_text(family = "Helvetica", size = (15), colour = "pink"),
                  legend.title = element_text(family = "Helvetica", size = (15), colour = "green",face="italic"),
                  legend.position="left")
  
a + mynamestheme 
```

### Does the correlation between gdp and fertility get stronger or weaker through time?


1. filter year to only values for 1800,1850, 1900, 1950, and 2000

2. filter out values without a continent entry

3. make a scatter plot with x axis = gdp_per_capita and y axis = fertility, colour based on continent (log scale the x-axis, use a discrete color scale)

4. facet by year

```{r}

```

### Animations with gganimate

https://github.com/thomasp85/gganimate

```{r}
library(gganimate)

country_colors <- readRDS("data/gapminder/country_colours.rds")

data %>% 
  filter(!is.na(continent)) %>% 
  ggplot(aes(gdp_per_capita, life_expectancy, size = population, colour = country)) +
  theme_classic() +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  ease_aes('linear')
```


# Reproducible research (Dan)


# Reproducible research with Rmarkdown

**Exercises:** Discuss with your partner 

- what is meant by Reproducible research?
- what can you do do make your research reproducible?
- who benefits from reproducible research?

## Keys to reproducible research

