---
title: "Data Manipulation and Visualization"
author: "Instructors: Daniel Falster, Will Cornwell, Dony Indiarto"
date: "Feb 2020"
output: html_document
---

# Introduction to using Rmarkdown (Rmd)


- File extension `*.Rmd`
- Can be notebook or markdown
- Includes text and code -> good for note taking 

Code chunks:

- typical R code
- are embedded in wrapper

```{r}

```

Text written in markdown

- Headers: `#`
- Bold: `**bold**`
- links: `[text](link)`

Hotcake short-cuts:

- in the top right of code

# Introduction to Movies data (Dan

```{r}
data <- read_csv("data/movie_profit/movie_profit.csv")
```

What variables?
```{r}
names(data)
```

An overview of the data

```{r}
View(data)
```

or summarise using the `skimr` package:

```{r}
skimr::skim(data)
```


# What is the tidyverse? (Dan)

* The [Tidyverse](http://tidyverse.org) is the name given to suite of R packages designed for seamless data analysis
* Designed to help you fall into a **"Pit of Success"**
* Tools designed for: 1) Turning data into tidy data, and 2) Plotting & analysing Tidy Data
* Dataframes (tibbles) are the universal "tidy" input and output
* Not one but a collection packages

Load individually or all together

```{r, eval = FALSE}
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
```
or
```{r, , eval = FALSE}}
library(tidyverse)
```

## `tibble` - A better data frame 


* Prints a short summary at console
* Generates suitable warnings 
* Won't convert strings to factors
* No partial column-name matching
* Behaves consistently with `[` and `[[`
* [tibble.tidyverse.org](tibble.tidyverse.org)

```{r}
data
```

# Reading data with `readr` (Dan)

Where possible, we recommend using 

- `csv` files (common separated files) to store data
- the `readr` package to load data

Why `readr`?

* Fast
* Won't convert strings to factors
* Can control how columns are parsed

```{r}
data <- read_csv("data/movie_profit/movie_profit.csv")
```

Has a bunch of extra arguments:

- guess_max
- trim_ws
- skip
- na

Tries to guess column types. Can override this by specifying exactly:


All characters
```{r}

read_csv("data/movie_profit/movie_profit.csv", 
         col_types = cols(.default = col_character())
)
```

Specify by variable:

```{r}
read_csv("data/movie_profit/movie_profit.csv", 
         col_types = cols(
           X1 = col_integer(),
           release_date = col_character(),
           movie = col_character(),
           production_budget = col_double(),
           domestic_gross = col_double(),
           worldwide_gross = col_double(),
           distributor = col_character(),
           mpaa_rating = col_character(),
           genre = col_character()
         )
)
```

Also other variants

- `read_tsv`: load tab-separated data
- `read_delim`: load file by specifying the delimiter

```{r}
read_delim("data/movie_profit/movie_profit.csv", ",")
```

To load data from excel spreadsheets, try the `readxl` package.

## Exercises

1. Use `read_csv` to load data file `data/plant_height/plantHeightSingleSpp.csv`

2. Use `read_csv` to load data file `data/sydney_beaches/temperature.csv`


# Data manipulation with `dplyr` (Dan)

Motivation:

- Data is never organized in the way you want it 
- High % of project is data wrangling
- Many many many modern jobs are data wrangling

Key ideas:

- Data is a noun
- Actions on data are verbs  

Main actions are verbs describing effect on data

- `select` -> subset columns 
- `filter` -> subset rows 
- `arrange`  –> order rows
- `rename`    –> rename variables
- `mutate`  –> make new variables
- `summarise`  –> summarise data
- `distinct` -> filter to each unique row


**Exercises:**

## Pipes


## Use "the pipe" [%>%](http://magrittr.tidyverse.org/reference/pipe.html) to connect expressions

* `%>%` "pipes" the **output** of the last expression as the **first input** of the next expression
* If you use RStudio, you can type the pipe with Ctrl + Shift + M if you have a PC or Cmd + Shift + M if you have a Mac.
* `%>%` is an **infix operator** -> expects commands on left & right
* Comes from the [magrittr](http://magrittr.tidyverse.org/reference/pipe.html) package


```{r}
data$distributor %>% unique()
```

```{r}
data$distributor %>% unique() %>% length()
data$distributor %>% n_distinct()
```

But you can control the input position of the next function with `.`:
```{r}
20 %>% seq(1, 4, length.out = .)
```

## Use pipes to join data verbs to make a data sentence

Tidyverse functions are written to work with pipes, i.e. most take the data as the first argument.

```{r}
data %>% filter(distributor == "Universal")
```
is the same as

```{r}
filter(data, distributor == "Universal")
```



```{r}
data %>%
  filter(distributor == "Universal") %>%
  select(movie, mpaa_rating)
```


**Exercises:** Apply the `dplyr` package and your new data wrangling skills to the movies dataset to 


1. Create a subset of the dataset consisting of only movies with budget < XXX
2. As above and only variables `XXX`, `XXX`, and `Longitude`
3. As above but with only one record per XXXX
3. As above but with data sorted by `XXXX`
4. Mutate a column to calculate XXX

# Imagine your plot (Will)

# Intro to data visualisation with `ggplot2` (Dony)

See slides in "slides/XXXX"

## Exercises

# Introduction to gapminder data (Dan)



# Tidy Data concept (Dony)

- What is tidy data?
- Converting from long to wide data
- see slides in XXXX

# Advanced data manipulation with `dplyr` (Dan)

- join
- group_by (summarise, mutate), 


**Exercise**:

Now using join and pipes, we can built a big table of data

```{r}

data <- 
  read_csv("data/gapminder/children_per_woman_total_fertility.csv") %>%   
  pivot_longer(-country, names_to = "year", values_to = "fertility") %>%
  # Add in population
  left_join(by=c("country", "year"),
            read_csv("data/gapminder/population_total.csv") %>%
              pivot_longer(-country, names_to = "year", values_to = "population")
  ) %>%
  # Add in gdp per capita
  left_join(by=c("country", "year"),
            read_csv("data/gapminder/income_per_person_gdppercapita_ppp_inflation_adjusted.csv") %>%   
              pivot_longer(-country, names_to = "year", values_to = "gdp_per_capita")
  ) %>%
  # Add in life expectancy
  left_join(by=c("country", "year"),
            read_csv("data/gapminder/life_expectancy_years.csv") %>%
              pivot_longer(-country, names_to = "year", values_to = "life_expectancy")
  ) %>%
  # Add in life continent
  left_join(by="country",
    read_csv("data/gapminder/continents.csv")
  ) %>% 
  # reorganise data
  select(continent, country, year, everything()) %>% 
  arrange(continent, country, year)
```

```{r}
 
```


# Advanced data visualisation with `ggplot2` (Will)

This session will show you some more advanaced plotting techniques, building on the basic plots from Day 1


## Facets

## Adjusting labels, sclaes, paletters

Is now achieved via the `scales` packages

```{r}
library(scales)
```


## Adding fitted lines

Can use lines `geom_smooth` to fit lines


## Multiple plot layouts with `patchwork`

## Themes


Can add a theme onto plot

```{r}

+ theme_classic()
```

Can also set default theme

```{r}
theme_set()
```

# Data wrangling & visualisation challenge (Dan)


# ggplot in talks (Dan)
	
# ggplot extensions (Will)
	
Base R vs ggplot

Other packages that build on ggplot

- hexbinplot
- ggbeeswarm
- plotly



# Reproducible research (Dan)


# Reproducible research with Rmarkdown

**Exercises:**

Discuss with your partner 

- what is meant by Reproducible research?
- what can you do do make your research reproducible?
- who benefits from reproducible research?

## Keys to reproducible research

