---
title: "Data Manipulation and Visualization (2)"
author: "Instructors: Daniel Falster, Will Cornwell, Dony Indiarto"
date: "Feb 2020"
output: html_document
---


# Introduction to gapminder data (Dan)

For day 2's exercises, we're going to use the famous gapminder dataset, maintained at [gapminder.org](http://gapminder.org)

**Exercise**: With your partner, review the `README` for the data at `data/gapminder/readme.md` to see the variables included. 


# Tidy Data concept (Dony)

- What is tidy data?
- Converting from long to wide data
- see slides in `slides/tidy_data_and_pivot_tables.pdf`

Read and inspect the structure of `life_expectancy_years.csv`

```{r}
library(tidyverse)
le <- read_csv("data/gapminder/life_expectancy_years.csv") # read the worldwide life expectancy data

skimr::skim(le) #summarise using the `skimr

View(le) #An overview of the data
```


Make the 'life expectancy' data tidy with `dplyr::pivot_longer( )`

```{r}
le_long<- 
  pivot_longer(data = le, # the data
               
                      # pivot all column into longer format except 'country'
                      cols=-country, 
               
                      #the name of the column made from the data stored in the col.names
                      names_to="year", 
                  
                      #the name of the column made from the data stored in the cells
                      values_to = "life_exp") 

View(le_long)
```

Make a figure of life expectancy trend over time

```{r}
 ggplot(le_long,                                   #data
 aes(x = year, y = life_exp, group = country)) +   #aesthetics
 geom_line() +	                                   #type of plot
 theme(legend.position ="none")                    #remove legend

```

Make the long 'life expectancy' data back to its original form with `dplyr::pivot_wider( )`

```{r}
le_wide <- pivot_wider(data = le_long,           #the data
                        names_from = "year",     #column to get the column names from
                        values_from = "life_exp")#column to get the values from

View(le_wide) 
```

**Exercises**:

Make these data tidy:

- `data\gapminder\children_per_woman_total_fertility.csv`
- `data\gapminder\income_per_person_gdppercapita_ppp_inflation_adjusted.csv`


# Advanced data manipulation with `dplyr` (Dan)

We're now going to focus in on some more advanced features of dplyr. First, we need to load a series of tables and use `pivot_longer` to reorgaise them. 

```{r}
# Fertility
data_fertility <- read_csv("data/gapminder/children_per_woman_total_fertility.csv") %>%   
  pivot_longer(-country, names_to = "year", values_to = "fertility")
  
# Population
data_population <- read_csv("data/gapminder/population_total.csv") %>%
  pivot_longer(-country, names_to = "year", values_to = "population")

# gdp per capita
data_gdp <- read_csv("data/gapminder/income_per_person_gdppercapita_ppp_inflation_adjusted.csv") %>%   
  pivot_longer(-country, names_to = "year", values_to = "gdp_per_capita")

# life expectancy
data_life_exp <- read_csv("data/gapminder/life_expectancy_years.csv") %>%
  pivot_longer(-country, names_to = "year", values_to = "life_expectancy")

# continents
data_continent <-  read_csv("data/gapminder/continents.csv")
  
```

But wouldn't it be nice to merge these into a single dataset?

That's where `dplyr`s collection of `join` functions come in  

## Joining tables

https://dplyr.tidyverse.org/reference/join.html

There's three broad types of joins

1. Mutating joins combine variables from the two data.frames
2. Filtering joins keep cases from the left-hand data.frame
3. Nesting joins create a list column of data.frames

We're going to focus here on one type of Mutating join:

- `left_join()`: return all rows from x, and all columns from x and y. Rows in x with no match in y will have NA values in the new columns. If there are multiple matches between x and y, all combinations of the matches are returned.

The key to making a join work is to indetify the variable from each table that link the tables together. 

Let's say we want add infomration about each countried continent into `data_population`

```{r}
left_join(data_population, data_continent, by="country")
```

Note that

- the order of tables matters
- it still works without the "by" argument, because it guesses which columns are in common
- you don't need to have the same named columns but this makes life much easier.

We can also join by two variables 

```{r}
left_join(data_population, data_fertility)
```

or

```{r}
left_join(data_population, data_fertility, by = c("country", "year"))
```

**Exercise:** 

1. Join the two data frames `data_population` and `data_life_exp`
2. Join the three data frames `data_population`, `data_life_exp` and `data_continent`


----

Now using join and pipes, we can built a big table of data containing all variables

```{r}

data <- 
  read_csv("data/gapminder/children_per_woman_total_fertility.csv") %>%   
  pivot_longer(-country, names_to = "year", values_to = "fertility") %>%
  # Add in population
  left_join(
    read_csv("data/gapminder/population_total.csv") %>%
              pivot_longer(-country, names_to = "year", values_to = "population")
  ) %>%
  # Add in gdp per capita
  left_join(
    read_csv("data/gapminder/income_per_person_gdppercapita_ppp_inflation_adjusted.csv") %>%   
              pivot_longer(-country, names_to = "year", values_to = "gdp_per_capita")
  ) %>%
  # Add in life expectancy
  left_join(
    read_csv("data/gapminder/life_expectancy_years.csv") %>%
              pivot_longer(-country, names_to = "year", values_to = "life_expectancy")
  ) %>%
  # Add in life continent
  left_join(
    read_csv("data/gapminder/continents.csv")
  ) %>% 
  # reorganise data
  select(continent, country, year, everything()) %>% 
  arrange(continent, country, year) %>% 
  mutate(year = as.integer(year))
```

What about Na's?

```{r}
filter(data, is.na(continent))
filter(data, is.na(continent)) %>% pull(country) %>% unique()

```

## Group_by (summarise, mutate)


## What to do when your big pipe thingo doesn't work?


# Advanced data visualisation with `ggplot2` (Will)

This session will show you some more advanced plotting techniques, building on the basic plots from Day 1 and using the gapminder data


## Adjusting labels, scales, palettes

Is now achieved via the `scales` packages

```{r}
library(scales)
```


```{r}
data %>% 
  filter(year==1900) %>%
  ggplot(aes(gdp_per_capita, life_expectancy, colour = continent, size=sqrt(population))) + 
  geom_point() + 
  scale_x_log10() + scale_y_log10() +
  labs(x = "GDP per capita ($)", y = "life expectancy (yr)") + 
  theme_classic()
```

## Facets


```{r}
data %>% 
  filter(year %in% seq(1800, 2020, by =20)) %>%
  ggplot(aes(gdp_per_capita, life_expectancy, colour = continent, size=population)) + 
  theme_classic()+
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_x_log10() + scale_y_log10() +
  scale_size(range = c(2, 12)) +
  labs(x = "GDP per capita ($)", y = "life expectancy (yr)") + 
  theme_classic() +
  facet_wrap(~year)
```

## Themes


Can add a theme onto plot

```{r}

+ theme_classic()
```

Can also set default theme

```{r}
theme_set()
```


## Adding fitted lines

Can use lines `geom_smooth` to fit lines


## Multiple plot layouts with `patchwork`



--------

# Data wrangling & visualisation challenge (Dan)


# ggplot in talks (Dan)
	
# ggplot extensions (Will)
	
Base R vs ggplot

Other packages that build on ggplot

- hexbinplot
- ggbeeswarm
- plotly
- gganimate


### Animations with gganimate

https://github.com/thomasp85/gganimate

```{r}
library(gganimate)

country_colors <- readRDS("data/gapminder/country_colours.rds")

data %>% 
  filter(!is.na(continent)) %>% 
  ggplot(aes(gdp_per_capita, life_expectancy, size = population, colour = country)) +
  theme_classic() +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  ease_aes('linear')
```


# Reproducible research (Dan)


# Reproducible research with Rmarkdown

**Exercises:** Discuss with your partner 

- what is meant by Reproducible research?
- what can you do do make your research reproducible?
- who benefits from reproducible research?

## Keys to reproducible research

