---
title: "Introduction to the Tidyverse"
author: "Daniel Falster"
date: "6 March 2018"
output: 
    ioslides_presentation
---

## What is the tidyverse

* The [Tidyverse](http://tidyverse.org) is the name given to suite of R packages designed for seamless data analysis
* Designed to help you fall into a **"Pit of Success"**
* Tools designed for: 1) Turning data into tidy data, and 2) Plotting & analysing Tidy Data
* Dataframes (tibbles) are the universal "tidy" input and output
* A tidy dataframe has: 1 observation per row, 1 variable per column
* Not one but a collection packages

## Load individually or all together

```{r, eval = FALSE}
library(dplyr)
library(readr)
library(tidyr)
...
```
or
```{r, results='hide'}
library(tidyverse)
```

## [`readr`](readr.tidyverse.org)- For reading and writing data

* Fast
* Won't convert strings to factors
* Can control how columns are parsed

```{r}
data <- read_csv("data/plant_height/plantHeight5Spp.csv")  
```

## [`tibble`](tibble.tidyverse.org) - A better data frame 

* Prints a short summary at console
* Generates suitable warnings 
* Won't convert strings to factors
* No partial column-name matching
* Behaves consistently with `[` and `[[`

```{r}
data
```

## [`dplyr`](dplyr.tidyverse.org) - For data wrangling

Main actions are verbs describing effect on data

- select  and  filter –> subset columns and rows 
- summarise  –> summarise data
- arrange  –> order rows
- rename    –> rename variables
- mutate  –> make new variables

## Use [select](http://dplyr.tidyverse.org/reference/select.html) to select variables

```{r}
select(data, site, Country, Family, Genus_species, height)
```


## Use [filter](http://dplyr.tidyverse.org/reference/filter.html) to select rows

```{r}
filter(data, Country == "Australia")
filter(data, Country == "Australia", Family == "Myrtaceae", height <3)
```

## Use [arrange](http://dplyr.tidyverse.org/reference/arrange.html) to sort by values in a column

```{r}
arrange(data, Country, Genus_species)
arrange(data, Country, Family, Genus_species)
```

## Use [rename](http://dplyr.tidyverse.org/reference/select.html) to rename a variable

```{r}
rename(data,  species = Genus_species)
```

## Use [mutate](http://dplyr.tidyverse.org/reference/mutate.html) creates new variables

```{r}
mutate(data,  log_ht = log(height))
```

## Use [summarise](http://dplyr.tidyverse.org/reference/summarise.html) for summary statistics

```{r}
summarise(data,  ht_mn = mean(height), ht_sd = sd(height), n = n())
```

## Use "the pipe" [%>%](http://magrittr.tidyverse.org/reference/pipe.html) to connect expressions

* `%>%` is an **infix operator** -> expects commands on left & right
* `%>%` "pipes" the **output** of the last expression as the **first input** of the next expression
* From [magrittr](http://magrittr.tidyverse.org/reference/pipe.html) package

```{r, eval = FALSE}
data$height %>% mean()
```

You can control the input position of the next function with `.`:
```{r}
20 %>% seq(1,4, length.out = .)
```

## Use pipes to join data verbs to make a data sentence

```{r, eval = FALSE}
data %>% 
  filter(Country == "Australia") %>% 
  select(Family, Genus_species, height) %>%
  mutate(log_height = log(height))
```

vs old-fashioned R
```{r, eval = FALSE}
data[data$Country == "Australia", c("Family", "Genus_species", "height")]
```

## Use [group_by](http://dplyr.tidyverse.org/reference/group_by.html) to run operations on subsets of data

On whole dataset
```{r, eval = FALSE}
data %>%
  summarise(ht_mn = mean(height), n = n())
```

Split by country
```{r}
data %>%
  group_by(Country) %>% 
  summarise(ht_mn = mean(height), ht_sd = sd(height), n = n())
```

## Use [count](http://dplyr.tidyverse.org/reference/count.html) to examine a single column

```{r}
data %>%
  group_by(Country) %>%
  count(Family)
```  

## Use [summarise_at](http://dplyr.tidyverse.org/reference/summarise_all.html) to apply a series of functions to a series of variables

```{r}
data %>%
  group_by(Country) %>% 
  summarise_at(vars(height, NPP), funs(mean, sd, length))
```

## The package [skimr](https://github.com/ropenscilabs/skimr) also provides nice summaries

```{r}
data %>%
  select_if(is.numeric) %>%
  skimr::skim()
```  

## Use [join](http://dplyr.tidyverse.org/reference/join.html) to connect two data frames

```{r}
band_members %>% left_join(band_instruments)
```

## Use tidyr's [gather](http://tidyr.tidyverse.org/reference/gather.html) to collapses multiple columns into key-value pairs

```{r}
stocks <- tibble(
  time = as.Date('2009-01-01') + 0:9,
  X = rnorm(10, 0, 1),
  Y = rnorm(10, 0, 2),
  Z = rnorm(10, 0, 4)
)

gather(stocks, stock, price, -time)
```


## Use tidyr's [spread](http://tidyr.tidyverse.org/reference/gather.html) to spread a key-value pair across multiple columns

i.e. opposite of `gather` 

```{r}
stocksm <- stocks %>% gather(stock, price, -time)
stocksm %>% spread(stock, price)
```


## Further information

- [Cheat sheets from Rstudio](https://www.rstudio.com/resources/cheatsheets/)
- [Tidyverse refernce materials](http://tidyr.tidyverse.org/)

Talks:

- [Hadley Wickham's "Data Science with R"](https://www.youtube.com/watch?v=K-ss_ag2k9E)
- [Emily Robinson's "Lesser known stars of the tidyverse"](https://www.rstudio.com/resources/videos/the-lesser-known-stars-of-the-tidyverse/)

Tutorials:

- [UNSW's environmental computing](http://environmentalcomputing.net/subsetting-data/)
- [Wickam's R for Data science](http://r4ds.had.co.nz/transform.html)
- [Stat 545 Into to dplyr](http://stat545.com/block009_dplyr-intro.html)
- [Data Camp's Into to tidyverse](http://varianceexplained.org/r/intro-tidyverse/)
- [Miles McBain's Into to tidyverse](https://github.com/MilesMcBain/tidycourse)

